---
title: "Gamma Processes"
author: "Devon Grossett"
format: 
  html:
    css: styles.css
    theme: cosmo
    code-fold: false
bibliography: Hawkes_Bibtex.bib
---

## Dirichlet Process

::: {#def-dirichlet_process .definition}
## Dirichlet Process -- DP [@Muller2015_BayesianNonparametrics])

Let $\alpha > 0$ and $G_0$ be a probability measure defined on $S$. A DP with 
parameters $(\alpha, G_0)$ is a random probability measure $G$ defined on $S$
which assigns probability $G(B)$ to every (measurable) set $B$ such that for 
each (measurable) finite partition $\{B_1,\ldots,B_k\}$ of $S$, the joint 
distribution of the vector $(G(B_1),\ldots,G(B_k))$ is the Dirichlet 
distribution with parameters
$$
(\alpha G_0(B_1),\ldots,\alpha G_0(B_k))
$$
:::

The process is usually denoted as $\mathrm{DP}(\alpha G_0)$, or 
$\mathrm{DP}(\alpha, G_0)$. The parameter $\alpha$ is called the precision or 
total mass parameter and $G_0$ is the centering measure.

**Stick Breaking Construction**  A useful constructive definition of a DP 
random probability measure is given by Sethurman (1994) and based on the 
discrete nature of the process $G(\cdot) = \sum_{k=1}^\infty \tilde{w}_k \delta_{\theta_k}(\cdot)$.
Formally, let $\tilde{w}_k = v_k \prod_{\ell<k}(1-v_{\ell})$ with 
$v_{\ell}\stackrel{iid}{\sim} \mathrm{Beta}(1, \alpha)$ and $\theta_k\stackrel{iid}{\sim} G_0$, 
where $\{v_k\}$ and $\{\theta_k\}$ are independent. Then 
$$
G(\cdot) = \sum_{k=1}^\infty \tilde{w}_k\delta_{\theta_k}(\cdot)
$$
defines a $\mathrm{DP}(\alpha G_0)$ random measure. 

In reality, we can choose some $k=K$ at which to truncate the series where 
beyond it, the values of $\tilde{w}_k$ become negligible. At that point, we can
set $v_K=1$, i.e. no stick breaking beyond this point.

```{r}
dp_stickbreak <- function(alpha, g0 = rexp, k_trunc = 100, ...) {
  # draw atoms from G_0
  theta <- g0(k_trunc, ...)
  
  # i.i.d. beta draws of v_k, truncate and set final value to 1
  v <- c(rbeta(k_trunc - 1, 1, alpha), 1)
  # stick remaining before each break
  remaining <- cumprod(c(1, 1 - v[-k_trunc]))
  # weights, with sum to one constraint
  w <- v*remaining
  w[k_trunc] <- 1 - sum(w[-k_trunc])
  
  return(list(theta = theta, weights = w, k_trunc = k_trunc))
}
```

An example with the base probability measure $H_0 \sim N(0,1)$, to visualise 
what the draws of a DP look like with varying $\alpha$.

```{r}
set.seed(1234)
par(mfrow = c(2, 2), mar = c(4, 5, 3, 1), oma = c(0, 0, 2, 0))

alphas <- c(1, 5, 20, 100)
truncs <- c(100, 200, 500, 2000) 

for (j in 1:4) {
  a <- alphas[j]
  k <- truncs[j]
  
  plot(NULL, xlim = c(-4, 4), ylim = c(0, 1),
       xlab = expression(t), ylab = expression(G(group("[", list(0, t), "]"))),
       main = bquote(alpha == .(a)))
  
  # add the theoretical Normal CDF (pnorm) in red
  curve(pnorm(x), from = -4, to = 4, col = "red", lwd = 2, add = TRUE)
  
  # draw 50 DP realisations
  for (i in 1:50) {
    # Using rnorm for the base measure g0
    dp <- dp_stickbreak(alpha = a, g0 = rnorm, k_trunc = k)
    ord <- order(dp$theta)
    lines(dp$theta[ord], cumsum(dp$weights[ord]), 
          type = "s", col = rgb(0, 0, 0, 0.1))
  }
}

# Add an overall title to the 2x2 grid
mtext(expression(paste("50 DP Realizations with ", G[0], " ~ N(0,1)")), 
      outer = TRUE, cex = 1.2, line = 0)
```

The Dirichlet Process is conjugate with respect to the sampling model 
$$
y_i \mid G \stackrel{iid}{\sim} G,\quad i=1,\ldots,n
$$
such that if $G \sim \mathrm{DP}(\alpha G_0)$, then
$$
G \mid y_1, \ldots, y_n \sim \mathrm{DP}\left(\alpha+n, 
  \frac{\alpha G_0 + \sum_{i=1}^n \delta_{y_i}}{\alpha + n}\right)
$$

```{r}
# function to sample from the posterior based on a DP(alpha, g0) prior
dp_post_stickbreak <- function(alpha, g0 = rexp, k_trunc = 100, obs, ...) {
  # helper function for simulation from centering measure of posterior DP
  rgn <- function(n_samples = 1, obs, alpha0, g00 = rnorm, ...) {
    prob_new <- alpha0/(alpha0 + length(obs))
    samples <- replicate(n_samples, {
      if (runif(1) < prob_new) {
        g00(1, ...)
      } else {
        sample(obs, 1)
      }
    })
  }
  dp_stickbreak(alpha = alpha + length(obs), g0 = rgn, k_trunc = k_trunc, 
                alpha0 = alpha, g00 = g0, obs = obs)
}
```

Lets imagine we observe some data from a mixture of normals 
$Y_i \sim 0.5N_1 + 0.5N_2$, $i=1,\ldots,n$ with 
$\mu_1=-1,\mu_2=1,\sigma_1^2=\sigma_2^2=0.5$ and $n=100$.

```{r}
set.seed(12345)

p <- rbinom(100, 1, 0.5)
y <- p*rnorm(100, -1, 0.5) + (1 - p)*rnorm(100, 1, 0.5)
grid_x <- seq(-4, 4, length.out = 500)
par(mfrow = c(2, 2), mar = c(4, 5, 3, 1), oma = c(0, 0, 2, 0))

for (j in 1:4) {
  a <- alphas[j]
  k <- truncs[j]
  
  plot(NULL, xlim = c(-4, 4), ylim = c(0, 1),
       xlab = expression(t), ylab = expression(G(group("[", list(0, t), "]"))),
       main = bquote(alpha == .(a)))
  
  all_cdfs <- matrix(0, nrow = 500, ncol = 50)
  
  # draw 50 posterior DP realisations
  for (i in 1:50) {
    dp <- dp_post_stickbreak(alpha = a, g0 = rnorm, k_trunc = k, obs = y)
    ord <- order(dp$theta)
    
    f_emp <- stepfun(dp$theta[ord], c(0, cumsum(dp$weights[ord])))
    all_cdfs[, i] <- f_emp(grid_x)
    
    lines(dp$theta[ord], cumsum(dp$weights[ord]), 
          type = "s", col = rgb(0, 0, 0, 0.1))
    
    legend("topleft", 
       legend = c("Posterior Mean", "True G"), 
       col = c("black", "red"), 
       lwd = c(1, 2), 
       # lty = c(1, 2),
       bty = "n",  
       cex = 0.8) 
  }
  
  curve(0.5*pnorm(x, -1, 0.5) + 0.5*pnorm(x, 1, 0.5), 
        from = -4, to = 4, col = "red", lwd = 2, add = TRUE)
  
  mean_cdf <- rowMeans(all_cdfs)
  lines(grid_x, mean_cdf, col = "black")
}

mtext(expression(paste("50 Posterior DP Realizations from prior DP(", alpha, ", N(0,1))")), 
      outer = TRUE, cex = 1.2, line = 0)
```

## Gamma Process Prior

::: {#def-gamma_process .definition}
## Gamma Process [@Phadia2016_PriorProcesses])

Let $\{Z_t\}_{t\geq0}$ be a stochastic process such that (i) $Z_0=0$, (ii) $Z_t$
has independent increments in non-overlapping intervals; and (iii) for $t>s$, 
the increment 
$$
(Z_t - Z_s) \sim \mathrm{Gamma}(c(\alpha(t)-\alpha(s)),c)
$$ 
where $c>0$ is a constant and $\alpha(t)$, $t>0$ be an increasing left-continuous 
function such that $\alpha(0)=0$. Then $Z_t$ is said to be a Gamma Process with 
parameters $c\alpha(t)$ and $c$ and denoted $\mathrm{GaP}(c\alpha(t),c)$.
:::

Thus we can view a Gamma Process as an increasing process with random 
increments, where the amount of increase over any interval follows a Gamma 
distribution. In Bayesian applications, the random non-negative measure $G$ on
$\mathbb{R}^+$ created by the Gamma Process (where $G([0,t])=Z_t$) can be used 
as a prior over non-negative finite measures, written as
$$
G \sim \mathrm{GaPP}(\alpha H_0,\beta)
$$
where we have reparameterised to have shape parameter $\alpha>0$, rate parameter 
$\beta > 0$, and a base probability measure over the space $\Theta=\mathbb{R}^+$.

A draw from a Gamma Process can be constructed from a DP draw by first drawing a 
random probability measure $\tilde{G} \sim \mathrm{DP}(\alpha H_0)$, and then 
scaling the resulting weights by a random total mass 
$\gamma \sim \mathrm{Gamma}(\alpha,\beta)$.

Using the stick breaking construction of the DP, this leads to the following
representation
$$
G(\cdot) = \gamma \sum_{k=1}^\infty v_k \prod_{\ell < k} (1-v_\ell) \delta_{\theta_k}(\cdot)
$$
where $\gamma \sim \mathrm{Gamma}(\alpha,\beta)$, 
$\theta_k \stackrel{iid}{\sim} H_0$, and
$v_k \stackrel{iid}{\sim} \mathrm{Beta}(1,\alpha)$ for $k=1,2,\ldots$. As above
for the DP, we truncate the weights at some $k=K$ sufficiently large so that 
this approximation is negligible.

```{r}
gp_from_dp <- function(alpha, beta, g0 = rexp, k_trunc = 100, ...) {
  gamma <- rgamma(1, alpha, rate = beta)
  dp <- dp_stickbreak(alpha = alpha, g0 = g0, k_trunc = k_trunc)
  return(list(
    theta = dp$theta, 
    weights = gamma*dp$weights, 
    gamma = gamma,
    k_trunc = dp$k_trunc
  ))
}
```

An example of draws of a Gamma process with base measure 
$H_0 \sim \mathrm{Exp}(1)$, where we have shown plots of the cumulative measure
$G([0,t])$

```{r}
set.seed(123456)
par(mfrow = c(2, 2), mar = c(4, 5, 3, 1), oma = c(0, 0, 2, 0))

alphas <- c(1, 5, 20, 50)
betas <- alphas
truncs <- c(100, 200, 500, 2000) 

for (j in 1:4) {
  a <- alphas[j]
  b <- betas[j]
  k <- truncs[j]
  
  plot(NULL, xlim = c(0, 5), ylim = c(0, 2),
       xlab = expression(t), ylab = expression(G(group("[", list(0, t), "]"))),
       main = bquote(alpha == .(a) ~","~ beta ==.(b)))
  
  # add the theoretical Normal CDF (pnorm) in red
  curve(pexp(x), from = 0, to = 5, col = "red", lwd = 2, add = TRUE)
  
  # draw 50 DP realisations
  for (i in 1:50) {
    # Using rnorm for the base measure g0
    gp <- gp_from_dp(alpha = a, beta = b, g0 = rexp, k_trunc = k)
    ord <- order(gp$theta)
    lines(gp$theta[ord], cumsum(gp$weights[ord]), 
          type = "s", col = rgb(0, 0, 0, 0.1))
  }
}
```

