---
title: "Gamma Processes"
author: "Devon Grossett"
format: 
  html:
    css: styles.css
    theme: cosmo
    code-fold: false
bibliography: Hawkes_Bibtex.bib
---

## Dirichlet Process

::: {#def-dirichlet_process .definition}
## Dirichlet Process -- DP [@Muller2015_BayesianNonparametrics]

Let $\alpha > 0$ and $G_0$ be a probability measure defined on $S$. A DP with 
parameters $(\alpha, G_0)$ is a random probability measure $G$ defined on $S$
which assigns probability $G(B)$ to every (measurable) set $B$ such that for 
each (measurable) finite partition $\{B_1,\ldots,B_k\}$ of $S$, the vector 
$(G(B_1),\ldots,G(B_k))$ has the joint distribution. 
$$
(G(B_1),\ldots,G(B_k)) \sim \mathrm{Dir}(\alpha G_0(B_1),\ldots,\alpha G_0(B_k))
$$
where $\mathrm{Dir}(\cdot)$ is the Dirichlet distribution.

:::

The process is usually denoted as $\mathrm{DP}(\alpha G_0)$, or 
$\mathrm{DP}(\alpha, G_0)$. The parameter $\alpha$ is called the precision or 
total mass parameter and $G_0$ is the centering measure.

**Stick Breaking Construction**  A useful constructive definition of a DP 
random probability measure is given by Sethurman (1994) and based on the 
discrete nature of the process $G(\cdot) = \sum_{k=1}^\infty \tilde{w}_k \delta_{\theta_k}(\cdot)$.
Formally, let $\tilde{w}_k = v_k \prod_{\ell<k}(1-v_{\ell})$ with 
$v_{\ell}\stackrel{iid}{\sim} \mathrm{Beta}(1, \alpha)$ and $\theta_k\stackrel{iid}{\sim} G_0$, 
where $\{v_k\}$ and $\{\theta_k\}$ are independent. Then 
$$
G(\cdot) = \sum_{k=1}^\infty \tilde{w}_k\delta_{\theta_k}(\cdot)
$$ {#eq-dp_stickbreak}
defines a $\mathrm{DP}(\alpha G_0)$ random measure. 

In reality, we can choose some $k=K$ at which to truncate the series where 
beyond it, the values of $\tilde{w}_k$ become negligible. At that point, we can
set $v_K=1$, i.e. no stick breaking beyond this point.

```{r dp_stickbreak_func}
dp_stickbreak <- function(alpha, g0 = rexp, k_trunc = 100, ...) {
  # draw atoms from G_0
  theta <- g0(k_trunc, ...)
  
  # i.i.d. beta draws of v_k, truncate and set final value to 1
  v <- c(rbeta(k_trunc - 1, 1, alpha), 1)
  # stick remaining before each break
  remaining <- cumprod(c(1, 1 - v[-k_trunc]))
  # weights, with sum to one constraint
  w <- v*remaining
  w[k_trunc] <- 1 - sum(w[-k_trunc])
  
  return(list(theta = theta, weights = w, k_trunc = k_trunc))
}
```

An example with the base probability measure $H_0 \sim N(0,1)$, to visualise 
what the draws of a DP look like with varying $\alpha$.

```{r plot_dp_draws}
set.seed(1234)
par(mfrow = c(2, 2), mar = c(4, 5, 3, 1), oma = c(0, 0, 2, 0))

alphas <- c(1, 5, 20, 100)
truncs <- c(100, 200, 500, 2000) 

for (j in 1:4) {
  a <- alphas[j]
  k <- truncs[j]
  
  plot(NULL, xlim = c(-4, 4), ylim = c(0, 1),
       xlab = expression(t), ylab = expression(G(group("[", list(0, t), "]"))),
       main = bquote(alpha == .(a)))
  
  # add the theoretical Normal CDF (pnorm) in red
  curve(pnorm(x), from = -4, to = 4, col = "red", lwd = 2, add = TRUE)
  
  # draw 50 DP realisations
  for (i in 1:50) {
    # Using rnorm for the base measure g0
    dp <- dp_stickbreak(alpha = a, g0 = rnorm, k_trunc = k)
    ord <- order(dp$theta)
    lines(dp$theta[ord], cumsum(dp$weights[ord]), 
          type = "s", col = rgb(0, 0, 0, 0.1))
  }
}

# Add an overall title to the 2x2 grid
mtext(expression(paste("50 DP Realizations with ", G[0], " ~ N(0,1)")), 
      outer = TRUE, cex = 1.2, line = 0)
```

The Dirichlet Process is conjugate with respect to the sampling model 
$$
y_i \mid G \stackrel{\rm iid}{\sim} G,\quad i=1,\ldots,n
$$
such that if $G \sim \mathrm{DP}(\alpha G_0)$, then
$$
G \mid y_1, \ldots, y_n \sim \mathrm{DP}\left(\alpha+n, 
  \frac{\alpha G_0 + \sum_{i=1}^n \delta_{y_i}}{\alpha + n}\right)
$$

```{r dp_posterior_draws}
# function to sample from the posterior based on a DP(alpha, g0) prior
dp_post_stickbreak <- function(alpha, g0 = rexp, k_trunc = 100, obs, ...) {
  # helper function for simulation from centering measure of posterior DP
  rgn <- function(n_samples = 1, obs, alpha0, g00 = rnorm, ...) {
    prob_new <- alpha0/(alpha0 + length(obs))
    samples <- replicate(n_samples, {
      if (runif(1) < prob_new) {
        g00(1, ...)
      } else {
        sample(obs, 1)
      }
    })
  }
  dp_stickbreak(alpha = alpha + length(obs), g0 = rgn, k_trunc = k_trunc, 
                alpha0 = alpha, g00 = g0, obs = obs)
}
```

Lets imagine we observe some data from a mixture of normals 
$Y_i \sim 0.5N_1 + 0.5N_2$, $i=1,\ldots,n$ with 
$\mu_1=-1,\mu_2=1,\sigma_1^2=\sigma_2^2=0.5$ and $n=100$.

```{r plot_dp_posterior_draws}
set.seed(12345)

p <- rbinom(100, 1, 0.5)
y <- p*rnorm(100, -1, 0.5) + (1 - p)*rnorm(100, 1, 0.5)
grid_x <- seq(-4, 4, length.out = 500)
par(mfrow = c(2, 2), mar = c(4, 5, 3, 1), oma = c(0, 0, 2, 0))

for (j in 1:4) {
  a <- alphas[j]
  k <- truncs[j]
  
  plot(NULL, xlim = c(-4, 4), ylim = c(0, 1),
       xlab = expression(t), ylab = expression(G(group("[", list(0, t), "]"))),
       main = bquote(alpha == .(a)))
  
  all_cdfs <- matrix(0, nrow = 500, ncol = 50)
  
  # draw 50 posterior DP realisations
  for (i in 1:50) {
    dp <- dp_post_stickbreak(alpha = a, g0 = rnorm, k_trunc = k, obs = y)
    ord <- order(dp$theta)
    
    f_emp <- stepfun(dp$theta[ord], c(0, cumsum(dp$weights[ord])))
    all_cdfs[, i] <- f_emp(grid_x)
    
    lines(dp$theta[ord], cumsum(dp$weights[ord]), 
          type = "s", col = rgb(0, 0, 0, 0.1))
    
    legend("topleft", 
       legend = c("Posterior Mean", "True G"), 
       col = c("black", "red"), 
       lwd = c(1, 2), 
       # lty = c(1, 2),
       bty = "n",  
       cex = 0.8) 
  }
  
  curve(0.5*pnorm(x, -1, 0.5) + 0.5*pnorm(x, 1, 0.5), 
        from = -4, to = 4, col = "red", lwd = 2, add = TRUE)
  
  mean_cdf <- rowMeans(all_cdfs)
  lines(grid_x, mean_cdf, col = "black")
}

mtext(expression(paste("50 Posterior DP Realizations from prior DP(", alpha, ", N(0,1))")), 
      outer = TRUE, cex = 1.2, line = 0)
```

## Gamma Process Prior

::: {#def-gamma_process .definition}
## Gamma Process [@Phadia2016_PriorProcesses]

Let $\{Z_t\}_{t\geq0}$ be a stochastic process such that (i) $Z_0=0$, (ii) $Z_t$
has independent increments in non-overlapping intervals; and (iii) for $t>s$, 
the increment 
$$
(Z_t - Z_s) \sim \mathrm{Gamma}(c(\alpha(t)-\alpha(s)),c)
$$ 
where $c>0$ is a constant and $\alpha(t)$, $t>0$ be an increasing left-continuous 
function such that $\alpha(0)=0$. Then $Z_t$ is said to be a Gamma Process with 
parameters $c\alpha(t)$ and $c$ and denoted $\mathrm{GaP}(c\alpha(t),c)$.
:::

Thus we can view a Gamma Process as an increasing process with random 
increments, where the amount of increase over any interval follows a Gamma 
distribution. In Bayesian applications, the random non-negative measure $G$ on
$\mathbb{R}^+$ created by the Gamma Process (where $G([0,t])=Z_t$) can be used 
as a prior over non-negative finite measures, written as
$$
G \sim \mathrm{GaPP}(\alpha H_0,\beta)
$$
where we have reparameterised to have shape parameter $\alpha>0$, rate parameter 
$\beta > 0$, and a base probability measure over the space $\Theta=\mathbb{R}^+$.

A draw from a Gamma Process can be constructed from a DP draw by first drawing a 
random probability measure $\tilde{G} \sim \mathrm{DP}(\alpha H_0)$, and then 
scaling the resulting weights by a random total mass 
$\gamma \sim \mathrm{Gamma}(\alpha,\beta)$.

Using the stick breaking construction of the DP, this leads to the following
representation:

$$
\begin{aligned}
\theta_k \mid H_0 &\stackrel{\rm iid}{\sim} H_0 \qquad k=1,2,\ldots\\
v_k \mid \alpha &\stackrel{\rm iid}{\sim} \mathrm{Be}(1,\alpha)\\
\tilde{w}_k &= v_k \prod_{\ell<k} (1-v_\ell)\\
\gamma &\sim \mathrm{Ga}(\alpha,\beta)\\
G({\rm d}u) &= \gamma\sum_{k=1}^\infty \tilde{w}_k \delta_{\theta_k}({\rm d}u)
\end{aligned}
$$ {#eq-gp_stickbreak}

Since the weights are stochastically decreasing in $k$, we can truncate our
draws of $v_k$ at some sufficiently large value of $K$, set the final weight 
$\tilde{w}_K$ to be one minus the sum of the previous $(1-K)$ weights,and 
represent the Gamma Process Prior as a finite sum with minimal impact on 
accuracy. This looks like:

$$
\begin{aligned}
\theta_k \mid H_0 &\stackrel{iid}{\sim} H_0 \qquad k=1,2,\ldots,K \\
v_k \mid \alpha &\stackrel{iid}{\sim} \mathrm{Be}(1,\alpha) \qquad k=1,2,\ldots,K-1 \\
\tilde{w}_k &= v_k \prod_{\ell < k} (1-v_\ell) \qquad k=1,2,\ldots,K-1 \\
\tilde{w}_K &= 1 - \sum_{k=1}^{K-1}\tilde{w}_k = \prod_{\ell=1}^{K-1}(1-v_\ell) \\
\gamma &\sim \mathrm{Ga}(\alpha,\beta) \\
G({\rm d}u) &= \gamma \sum_{k=1}^K \tilde{w}_k \delta_{\theta_k}({\rm d}u)
\end{aligned}
$$ {#eq-gp_stickbreak_truncate}

```{r gp_draws}
gp_from_dp <- function(alpha, beta, g0 = rexp, k_trunc = 100, ...) {
  gamma <- rgamma(1, alpha, rate = beta)
  dp <- dp_stickbreak(alpha = alpha, g0 = g0, k_trunc = k_trunc, ...)
  return(list(
    theta = dp$theta, 
    weights = gamma*dp$weights, 
    gamma = gamma,
    k_trunc = dp$k_trunc
  ))
}
```

An example of draws of a Gamma process with base measure 
$H_0 \sim \mathrm{Exp}(1)$, where we have shown plots of the cumulative measure
$G([0,t])$

```{r plot_gp_draws}
set.seed(123456)
par(mfrow = c(2, 2), mar = c(4, 5, 3, 1), oma = c(0, 0, 2, 0))

alphas <- c(1, 5, 20, 50)
betas <- alphas
truncs <- c(100, 200, 500, 2000) 

for (j in 1:4) {
  a <- alphas[j]
  b <- betas[j]
  k <- truncs[j]
  
  plot(NULL, xlim = c(0, 5), ylim = c(0, 2),
       xlab = expression(t), ylab = expression(G(group("[", list(0, t), "]"))),
       main = bquote(alpha == .(a) ~","~ beta ==.(b)))
  
  # add the theoretical Normal CDF (pnorm) in red
  curve(pexp(x), from = 0, to = 5, col = "red", lwd = 2, add = TRUE)
  
  # draw 50 DP realisations
  for (i in 1:50) {
    # Using rnorm for the base measure g0
    gp <- gp_from_dp(alpha = a, beta = b, g0 = rexp, k_trunc = k)
    ord <- order(gp$theta)
    lines(gp$theta[ord], cumsum(gp$weights[ord]), 
          type = "s", col = rgb(0, 0, 0, 0.1))
  }
}
```

## Hawkes Process with Gamaa Process Priors (GaPP)

For a univariate Hawkes Process, the conditional intensity function is:
$$
\lambda^*(t) = \lambda_0 + \sum_{t_i<t}\mu(t-t_i)
$$

We will consider a GaPP for $\mu(\cdot)$ constructed so that any realisation of
it is a non-increasing function (a similar constraint to the common exponential 
and power decay Hawkes kernels). 

We will define a non-parametric form of $\mu(\cdot)$ using a GaPP:
$$
\mu(t) = \int_t^\infty G({\rm d}u) = \gamma \sum_{k=1}^K\tilde{w}_k I(\theta_k > t)
$$

Following 
@Arnold2023_BayesNonparametricBathtub where they reparameterise the background 
intensity (or in their case, the baseline hazard rate) as 
$\lambda_0 = \gamma \eta$, the conditional intesnity function of the Hawkes 
Process can be written as
$$
\begin{aligned}
\lambda^*(t \mid \gamma, \eta, \boldsymbol{\theta}, \mathbf{v}) 
  &= \gamma \eta + \gamma \sum_{t_i<t}\sum_{k=1}^K\tilde{w}_k I(\theta_k > t-t_i) \\
  &= \gamma \tilde{\lambda}^*(t \mid \eta, \boldsymbol{\theta}, \mathbf{v})
\end{aligned}
$$ {#eq-hp_intensity_gapp}
where 
$$
\tilde{\lambda}^*(t \mid \eta, \boldsymbol{\theta}, \mathbf{v})
  = \eta + \sum_{t_i<t}\sum_{k=1}^K\tilde{w}_k I(\theta_k > t-t_i)
$$

For additional interpretability, we can swap the summation order and note that 
in @eq-hp_intensity_gapp, $\sum_{t_i<t}I(t-t_i < \theta_k)$ is simply counting 
the number of events before time $t$ that occured within the last $\theta_k$ 
time units. Therefore, the Gamma Process Prior for the Hawkes Process has the 
interpretation that each event that occured within the last $\theta_k$ time 
units of a given time $t$ contributes $\gamma\tilde{w}_k$ to the conditional 
intensity, and this effect stacks for all atoms $\theta_k$ drawn from the 
Gamma Process Prior. Based on this interpretation, we can define a counting 
process from that term of the conditional intensity
$$
N_k(t_i) = \sum_{j=1}^{i-1}I(t_i-t_j < \theta_k) = \#\{j<i:t_i-t_j<\theta_k\}
$$

Before we calculate the compensator of a Hawkes Process of this form, we note
that:
$$
\begin{aligned}
\int_0^t \mu(u){\rm d}u &= \int_0^t \gamma \sum_{k=1}^K\tilde{w}_k I(u < \theta_k) {\rm d}u \\
&= \gamma \sum_{k=1}^K \tilde{w}_k \int_0^t I(u < \theta_k) {\rm d}u \\
&= \gamma \sum_{k=1}^K \int_0^{\mathrm{min}\{t,\theta_k\}} {\rm d}u \\
&= \gamma \sum_{k=1}^K \tilde{w}_k \mathrm{min}\{t,\theta_k\}
\end{aligned}
$$

Therefore, as the compensator of a Hawkes Process is 
$\Lambda^*(t) = \int_0^t \lambda^*(u) {\rm d}u = \lambda_0 t + \sum_{i=1}^{N_t}  M(t - t_i)$,
where $M(t) = \int_0^t\mu(u) {\rm d}u$, we can write the compensator of a 
Hawkes Process with a truncated Gamma Process Prior as: 
$$
\begin{aligned}
\Lambda^*(T \mid \gamma, \eta, \boldsymbol{\theta}, \mathbf{v}) 
  &= \gamma \eta T + \gamma \sum_{i=1}^{N_T}\sum_{k=1}^K \tilde{w}_k \min\{T-t_i,\theta_k\} \\
  &= \gamma \eta T + \gamma \sum_{k=1}^K \tilde{w}_k \sum_{i=1}^{N_T} \min\{T-t_i,\theta_k\} \\
  &= \gamma \tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})
\end{aligned}
$$ {#eq-hp_compensator_gapp}
where 
$$
\tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})
  = \eta T +  \sum_{k=1}^K \tilde{w}_k \sum_{i=1}^{N_T} \min\{T-t_i,\theta_k\}
$$

The likelihood for a Hawkes Process observed over the period $[0,T]$ with a
Gamma Process Prior for the excitation function is then
$$
\begin{aligned}
L &= \left[ \prod_{i=1}^{N_T} \gamma\tilde{\lambda}^*(t_i \mid \eta, \boldsymbol{\theta}, \mathbf{v}) \right] 
    e^{-\gamma\tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})} \\
  &= \left[ \prod_{i=1}^{N_T} \gamma\left(\eta + \sum_{k=1}^K\tilde{w}_k N_k(t_i)\right)\right] \\
  &\quad\times\exp\left\{ -\gamma \left( \eta T +  \sum_{k=1}^K \tilde{w}_k \sum_{i=1}^{N_T} \min\{T-t_i,\theta_k\} \right) \right\} \\
\end{aligned}
$$ {#eq-hp_likelihood_gapp}
with log-likelihood
$$
\begin{aligned}
\ell
  &= N_T\log\gamma + \sum_{i=1}^{N_T} \log \left(\eta + \sum_{k=1}^K\tilde{w}_k N_k(t_i)\right) \\
  &\quad -\gamma \eta T + \gamma\sum_{k=1}^K \tilde{w}_k \sum_{i=1}^{N_T} \min\{T-t_i,\theta_k\} \\
\end{aligned}
$$ {#eq-hp_loglikelihood_gapp}

### Inference

Our full model with chosen priors is

$$
\begin{aligned}
t_i \mid \gamma, \eta, \boldsymbol{\theta}, \mathbf{v} 
  &\stackrel{\rm iid}{\sim} \mathrm{HP}(\gamma \tilde{\lambda}^*(\cdot \mid \eta, \boldsymbol{\theta}, \mathbf{v}))
  \qquad i=1,\ldots,N_T \\
  \theta_k \mid H_0, \phi &\stackrel{\rm iid}{\sim} \mathrm{Exp}(\phi) \qquad k=1,\ldots,K \\
  v_k \mid \alpha &\stackrel{\rm iid}{\sim} \mathrm{Beta}(1,\alpha) \qquad k=1,\ldots,K-1 \\
  \gamma \mid \alpha, \beta &\sim \mathrm{Ga}(\alpha, \beta) \\
  \eta \mid \nu &\sim \mathrm{Exp}(\nu) \\
  \alpha \mid a_1, a_2 &\sim \mathrm{Ga}(a_1, a_2) \\
  \beta \mid b_1, b_2 &\sim \mathrm{Ga}(b_1, b_2) \\
  \phi \mid f_1, f_2 &\sim \mathrm{Ga}(f_1, f_2) \\
\end{aligned}
$$
The following three parameters have simple Gibbs updates due to conjugacy:

#### $\gamma$:

$$
\begin{aligned}
p(\gamma \mid -) 
  &\propto p(\mathbf{t} \mid -)p(\gamma \mid \alpha,\beta) \\
  &\propto \prod_{i=1}^{N_T} \left[\gamma \tilde{\lambda}^*(t_i \mid \eta, \boldsymbol{\theta}, \mathbf{v})\right]
    e^{-\gamma\tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})}
    \gamma^{\alpha-1}e^{-\beta\gamma} \\
  &\propto \gamma^{N_T}e^{-\gamma\tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})} 
    \gamma^{\alpha-1}e^{-\beta\gamma} \\ 
  &\propto \gamma^{(\alpha + N_T)-1}e^{-\gamma(\beta + \tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v}))}
\end{aligned}
$$

Therefore the Gibbs update is 
$$
\gamma \mid - 
  \sim \mathrm{Ga}\left(\alpha + N_T,\beta + \tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})\right)
$$


#### $\beta$:

$$
\begin{aligned}
p(\beta \mid -) 
  &\propto p(\gamma \mid \alpha,\beta)p(\beta) \\
  &\propto \beta^{\alpha}e^{-\beta\gamma} \beta^{b_1-1}e^{-b_2\beta} \\
  &\propto \beta^{(\alpha + b_1)-1}e^{-\beta(\gamma + b_2)}
\end{aligned}
$$


Therefore the Gibbs update is 
$$
\beta \mid - 
  \sim \mathrm{Ga}\left( \alpha + b_1,\gamma + b_2 \right)
$$

#### $\phi$:

$$
\begin{aligned}
p(\phi \mid -) 
  &\propto p(\boldsymbol{\theta} \mid \phi)p(\phi) \\
  &\propto \left[\prod_{k=1}^K \phi e^{-\phi \theta_k}\right]\phi^{f_1-1}e^{-\phi f_2} \\
  &\propto \phi^K e^{-\phi\sum_{k=1}^K\theta_k}\phi^{f_1-1}e^{-\phi f_2} \\
  &\propto \phi^{(f_1 + K)-1}e^{-\phi\left(f_2 + \sum_{k=1}^K\theta_k\right)}
\end{aligned}
$$


Therefore the Gibbs update is 
$$
\phi \mid - 
  \sim \mathrm{Ga}\left( f_1 + K, f_2 + \sum_{k=1}^K\theta_k \right)
$$

The remaining parameters require Metropolis-Hastings updates.

$$$$ $\eta$:

$$
\begin{aligned}
p(\eta \mid -) 
  &\propto p(\mathbf{t} \mid -)p(\eta) \\
  &\propto \prod_{i=1}^{N_T} \left[\gamma \tilde{\lambda}^*(t_i \mid \eta, \boldsymbol{\theta}, \mathbf{v})\right]
    e^{-\gamma\tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})}
   e^{-\nu \eta} \\
  &\propto \prod_{i=1}^{N_T} \left[\tilde{\lambda}^*(t_i \mid \eta, \boldsymbol{\theta}, \mathbf{v})\right] 
    e^{-\gamma \eta T - \nu\eta}
\end{aligned}
$$

M-H update of $\tilde{\eta} = \log\eta$ with Normal proposal: 
$\mathrm{N}(\tilde{\eta},\sigma_\tilde{\eta}^2)$. This transformation has Jacobian 
$\left|\frac{{\rm d}\eta}{{\rm}d\tilde{\eta}}\right| = \eta$, and therefore
$$
\begin{aligned}
p(\tilde{\eta} \mid -) 
  &= p(\eta = e^{\tilde{\eta}} \mid -)\left|\frac{{\rm d}\eta}{{\rm}d\tilde{\eta}}\right| \\
  &= p(\eta = e^{\tilde{\eta}} \mid -) \eta \\
\log p(\tilde{\eta} \mid -) &= \log p(\eta = e^{\tilde{\eta}} \mid -) + \log \eta
\end{aligned}
$$

As the proposal distribution is symmetric, the acceptance ratio is simply 
$$
r = \frac{p(\tilde{\eta}^* \mid -)}{p(\tilde{\eta} \mid -)}
$$

$$$$ $\alpha$:

$$
\begin{aligned}
p(\alpha \mid -) 
  &\propto p(\gamma \mid \alpha, \beta)p(\mathbf{v} \mid \alpha)p(\alpha) \\
  &\propto \frac{\beta^\alpha}{\Gamma(\alpha)} \gamma^{\alpha-1}
    \left[ \prod_{k=1}^{K-1} \alpha (1-v_k)^{\alpha-1} \right] \alpha^{a_1 - 1}e^{-a_2\alpha} \\
  &\propto \frac{1}{\Gamma(\alpha)} \alpha^{a_1 + K - 2} (\beta\gamma)^\alpha \tilde{w}_K^\alpha e^{-a_2\alpha} \\
  &\propto \frac{1}{\Gamma(\alpha)} \alpha^{a_1 + K - 2} e^{-\alpha(-\log{\beta\gamma} - \log\tilde{w}_K + a_2)} \\
\end{aligned}
$$

As above, the M-H update is of $\tilde{\alpha} = \log\alpha$ with Normal proposal: 
$\mathrm{N}(\tilde{\alpha},\sigma_\tilde{\alpha}^2)$. 

As the proposal distribution is symmetric, the acceptance ratio is simply 
$$
r = \frac{p(\tilde{\alpha}^* \mid -)}{p(\tilde{\alpha} \mid -)}
$$

$$$$ $\theta_k$:

$$
\begin{aligned}
p(\theta_k \mid -) 
  &\propto p(\mathbf{t} \mid -) p(\theta_k \mid \phi) \\
  &\propto \left[\prod_{i=1}^{N_t}\gamma\tilde{\lambda}^*(t_i \mid \eta, \boldsymbol{\theta}, \mathbf{v})\right] 
    e^{-\gamma\tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})} e^{-\phi\theta_k} \\
  &\propto \left[\prod_{i=1}^{N_t}\tilde{\lambda}^*(t_i \mid \eta, \boldsymbol{\theta}, \mathbf{v})\right] 
    e^{-\gamma\tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})} e^{-\phi\theta_k} \\
\end{aligned}
$$

The sampling scheme for this comes from @Arnold2020_BayesNonparametricBathtub, 
where mixing is assisted by selecting an atom at random from the unscaled 
weights $\tilde{w}$. For each pass of the sampler we update $k_\text{sim} \leq K$
randomly selected atoms (selected with replacement). This ensures that the most
influential support points are updated with greatest frequency.

Of the selected atoms, we update $\tilde{\theta}_k = \log\theta_k$ with Normal
proposal: $\mathrm{N}(\tilde{\theta}_k,\sigma_\tilde{\theta}^2)$.

As the proposal distribution is symmetric, the acceptance ratio is simply 
$$
r = \frac{p(\tilde{\theta}_k^* \mid -)}{p(\tilde{\theta}_k \mid -)}
$$

$$$$ $v_k$:

$$
\begin{aligned}
p(v_k \mid -) 
  &\propto p(\mathbf{t} \mid -) p(v_k \mid \alpha) \\
  &\propto \left[\prod_{i=1}^{N_t}\gamma\tilde{\lambda}^*(t_i \mid \eta, \boldsymbol{\theta}, \mathbf{v})\right] 
    e^{-\gamma\tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})}  (1-v_k)^{\alpha-1} \\
  &\propto \left[\prod_{i=1}^{N_t}\tilde{\lambda}^*(t_i \mid \eta, \boldsymbol{\theta}, \mathbf{v})\right] 
    e^{-\gamma\tilde{\Lambda}^*(T \mid \eta, \boldsymbol{\theta}, \mathbf{v})} (1-v_k)^{\alpha-1} \\
\end{aligned}
$$

M-H update of $\tilde{v}_k = \text{logit}(\eta)$ with Normal proposal: 
$\mathrm{N}(\tilde{v}_k,\sigma_\tilde{v}^2)$. This transformation has Jacobian 
$\left|\frac{{\rm d}v_k}{{\rm}d\tilde{v}_k}\right| = v_k(1-v_k)$, and therefore
$$
\begin{aligned}
p(\tilde{v}_k \mid -) 
  &= p(v_k = \text{logit}^{-1}(\tilde{v}_k) \mid -)\left|\frac{{\rm d}v_k}{{\rm}d\tilde{v}_k}\right| \\
  &= p(v_k = \text{logit}^{-1}(\tilde{v}_k) \mid -) v_k(1-v_k) \\
\log p(\tilde{v}_k \mid -) &= \log p(v_k = \text{logit}^{-1}(\tilde{v}_k) \mid -) + \log v_k + \log(1-v_k)
\end{aligned}
$$

As the proposal distribution is symmetric, the acceptance ratio is simply 
$$
r = \frac{p(\tilde{v}_k^* \mid -)}{p(\tilde{v}_k \mid -)}
$$

### Implementation

```{r}
lambda_tilde <- function(t, eta, theta, w_raw, times) {
  diffs <- t - times
  N_k <- colSums(outer(diffs, theta, FUN = "<"))
  
  return(eta + sum(w_raw*N_k))
}

compensator_tilde <- function(t, eta, theta, w_raw, times) {
  diffs <- t - times
  min_matrix <- outer(diffs, theta, FUN = pmin)
  inner_sums <- colSums(min_matrix)
  return(eta*t + sum(w_raw*inner_sums))
}

logit <- function(x) {log(x/(1 - x))}
logit_inv <- function(x) {exp(x)/(1 + exp(x))}

## Gibbs proposals
sample_gamma <- function(T_max, alpha, beta, N_T, eta, theta, w_raw, times) {
  return(rgamma(
    1, 
    shape = alpha + N_T,
    rate = beta + compensator_tilde(T_max, eta, theta, w_raw, times)
  ))
}

sample_beta <- function(alpha, gamma, b_1, b_2) {
  return(rgamma(1, shape = alpha + b_2, rate = gamma + b_2))
}

sample_phi <- function(theta, f_1, f_2) {
  return(rgamma(1, shape = f_1 + length(theta), rate = f_2 + sum(theta)))
}

## M-H target distributions
logpost_alpha <- function(log_alpha, beta, gamma, w_raw, a_1, a_2) {
  alpha <- exp(log_alpha)
  K <- length(w_raw)
  w_K <- w_raw[K]
  
  return(
    log(1/gamma(alpha)) 
    + (a_1 + K - 2)*log(alpha) 
    + alpha*(a_2 - log(beta*gamma) - log(w_K))
    + log_alpha  #Jacobian
  )
}

logpost_eta <- function(log_eta, gamma, theta, w_raw, nu, T_max, times) {
  eta <- exp(log_eta)
  
  sum <- 0
  for (t_i in times) {
    sum = sum + log(lambda_tilde(t_i, eta, theta, w_raw, times[times < t_i]))
  }
  
  return(
    sum - eta*(gamma*T_max + nu)
    + log_eta  #Jacobian
  )
}

logpost_theta_k <- function(log_theta_k, k, gamma, eta, theta, w_raw, phi, T_max, times) {
  theta_k <- exp(log_theta_k)
  theta[k] <- theta_k
  
  sum <- 0
  for (t_i in times) {
    sum = sum + log(lambda_tilde(t_i, eta, theta, w_raw, times[times < t_i]))
  }
  
  return(
    sum 
    - gamma*compensator_tilde(T_max, eta, theta, w_raw, times)
    - phi*theta_k
    + log_theta_k  #Jacobian
  )
}

logpost_v_k <- function(logit_v_k, k, gamma, eta, theta, v, alpha, T_max, times) {
  v_k <- logit_inv(logit_v_k)
  v[k] <- v_k
  
  # reconstruct raw stick-break weights
  K <- length(theta)
  remaining <- cumprod(c(1, 1 - v))
  w_raw <- c(v, 1)*remaining
  w_raw[K] <- 1 - sum(w_raw[-K])
  
  sum <- 0
  for (t_i in times) {
    sum = sum + log(lambda_tilde(t_i, eta, theta, w_raw, times[times < t_i]))
  }
  
  return(
    sum 
    - gamma*compensator_tilde(T_max, eta, theta, w_raw, times)
    + (alpha - 1)*log(1 - v_k)
    + log(v_k) + log(1 - v_k)  #Jacobian (logit)
  )
}

## Function for random walk Metropolis-Hastings step
mh_step <- function(current, logpost_fun, proposal_sd, ...) {
  # Automatically get the name of the function passed in
  fun_name <- deparse(substitute(logpost_fun))
  
  # random walk sampler (in transformed space e.g. log, logit)
  proposal <- rnorm(1, current, proposal_sd)
  
  lp_prop <- logpost_fun(proposal, ...)
  lp_curr <- logpost_fun(current, ...)
  
  # Check for length zero or NA
  if (length(lp_prop) == 0 || is.na(lp_prop)) {
    stop(paste("Error in", fun_name, ": Proposal returned length 0 or NA. Value:", lp_prop))
  }
  if (length(lp_curr) == 0 || is.na(lp_curr)) {
    stop(paste("Error in", fun_name, ": Current returned length 0 or NA. Value:", lp_curr))
  }
  #log(acceptance ratio). the log-posterior functions will have necessary Jacobians
  log_acc <- lp_prop - lp_curr
  
  if (log(runif(1)) < log_acc) {
    return(list(value = proposal, accept = 1))
  } else {
    return(list(value = current, accept = 0))
  }
}

## Code for Metropolis-within-Gibbs sampler
run_sampler <- function(times, T_max, n_iter, init, prior_params, proposal_sds) {
  
  samples <- matrix(NA, n_iter, sum(lengths(init)))
  acceptance <- matrix(NA, n_iter, 7)
  colnames(samples) <- c("gamma", "beta", "phi", "alpha", "eta", 
                         paste0("theta_", 1:(length(init$theta))), 
                         paste0("v_", 1:(length(init$v))))
  colnames(acceptance) <- c("gamma", "beta", "phi", "alpha", "eta", "theta", "v")
  
  # initialise
  gamma <- init$gamma
  beta  <- init$beta
  phi   <- init$phi
  alpha <- init$alpha
  eta   <- init$eta
  theta <- init$theta
  v     <- init$v
  
  # some preliminary calculations
  N <- length(times)
  N_T <- times[N]
  K <- length(init$theta)
  
  pb <- txtProgressBar(min = 0,      
                       max = n_iter, 
                       style = 3,    
                       width = 50,   
                       char = "=")   
  
  for (iter in 1:n_iter) {
    
    # recalculate stick-break weights at the start of each iteration
    remaining <- cumprod(c(1, 1 - v))
    w_raw <- c(v, 1)*remaining
    w_raw[K] <- 1 - sum(w_raw[-K])
    
    # Gamma (Gibbs step)
    gamma <- sample_gamma(T_max, alpha, beta, N_T, eta, theta, w_raw, times)
    acceptance[iter, "gamma"] <- 1
    
    # Beta (Gibbs step)
    beta <- sample_beta(alpha, gamma, prior_params$b_1, prior_params$b_2)
    acceptance[iter, "beta"] <- 1
    
    # Phi (Gibbs step)
    phi <- sample_phi(theta, prior_params$f_1, prior_params$f_2)
    acceptance[iter, "phi"] <- 1
    
    # Alpha (Metropolis-within-Gibbs)
    res <- mh_step(log(alpha), logpost_alpha, proposal_sds$alpha,
                   beta = beta, gamma = gamma, w_raw = w_raw, 
                   a_1 = prior_params$a_1, a_2 = prior_params$a_2)
    alpha <- exp(res$value)
    acceptance[iter, "alpha"] <- res$accept
    
    # Eta (Metropolis-within-Gibbs)
    res <- mh_step(log(eta), logpost_eta, proposal_sds$eta,
                   gamma = gamma, theta = theta, w_raw = w_raw, 
                   nu = prior_params$nu, T_max = T_max, times = times)
    eta <- exp(res$value)
    acceptance[iter, "eta"] <- res$accept
    
    # Theta_k (Metropolis-within-Gibbs)
    k_theta <- sample(1:K, 1, prob = w_raw)
    res <- mh_step(log(theta[k_theta]), logpost_theta_k, proposal_sds$theta_k, 
                   k = k_theta, gamma = gamma, eta = eta, theta = theta, 
                   w_raw = w_raw, phi = phi, T_max = T_max, times = times)
    theta[k_theta] <- exp(res$value)
    acceptance[iter, "theta"] <- res$accept
    
    # v_k (Metropolis-within-Gibbs)
    k_v <- sample(1:(K-1), 1, prob = w_raw[-K]/(sum(w_raw[-K])))
    res <- mh_step(logit(v[k_v]), logpost_v_k, proposal_sds$v_k, k = k_v, 
                   gamma = gamma, eta = eta, theta = theta, v = v, 
                   alpha = alpha, T_max = T_max, times = times)
    v[k_v] <- logit_inv(res$value)
    acceptance[iter, "v"] <- res$accept
    
    
    samples[iter, ] <- c(gamma, beta, phi, alpha, eta, theta, v)
    setTxtProgressBar(pb, iter)
  }
  
  close(pb)
  acceptance_rates <- colMeans(acceptance)
  
  return(list(samples = samples, acceptance_rates = acceptance_rates))
}
```


```{r}
library(HawkesProject)


set.seed(6)
gp <- gp_from_dp(alpha = 2, beta = 1, g0 = rexp, k_trunc=40, rate = 2)
example_decay <- function(x) {
  rho <- 0.6
  thetas <- gp$theta
  weights <- gp$weights/sum(gp$theta*gp$weights)
  vapply(x, function(xi) rho*sum(weights[thetas > xi]), numeric(1))
}

x_grid <- seq(0, 2.5, length.out = 500)
y <- example_decay(x_grid)
plot(x_grid, y, type = "s",
     xlab = "x", ylab = "Decay",
     main = "Example Decay Function")

set.seed(456)
simulated_process <- sim_hp(T_max = 200, lambda = 0.5, mu = example_decay)
plot(simulated_process)
```

```{r}
prior_params <- list(a_1 = 1, a_2 = 1, f_1 = 1, f_2 = 1, r_1 = 4, r_2 = 4, nu = 0.1)
proposal_sds <- list(lambda0 = 0.2, rho = 1, theta_k = 2, v_k = 2)
n_iter <- 2000
set.seed(6789)
init_list <- list(lambda0 = 1, rho = 0.5, theta = runif(20, 0, 5), 
                  v = runif(19, 0, 1), alpha = 1, phi = 2)

chains <- run_sampler(simulated_process$events, simulated_process$T_max, 
                      n_iter, init_list, prior_params, proposal_sds)
```

```{r}
stickbreak <- function(v) {
  K <- length(v) + 1
  
  remaining <- cumprod(c(1, 1 - v))
  w_raw <- c(v, 1) * remaining
  
  # enforce sum to 1 numerically
  w_raw[K] <- 1 - sum(w_raw[-K])
  
  return(w_raw)
}

compute_w <- function(v, theta) {
  w_raw <- stickbreak(v)
  w <- w_raw / sum(w_raw * theta)
  return(w)
}

kernel_eval <- function(x, theta, w, rho) {
  sapply(x, function(xi) {
    rho * sum(w[theta > xi])
  })
}

samples <- chains$samples[-(1:100), ]
K <- 20
theta_cols <- paste0("theta", 1:K)
v_cols <- paste0("v", 1:(K-1))

x_grid <- seq(0, 5, length.out = 200)

n_draws <- nrow(samples)
kernel_draws_full <- matrix(0, nrow = n_draws, ncol = length(x_grid))
for (i in 1:n_draws) {
  s <- samples[i, ]
  
  theta <- as.numeric(s[theta_cols])
  v <- as.numeric(s[v_cols])
  rho <- s["rho"]
  
  w <- compute_w(v, theta)
  
  kernel_draws_full[i, ] <- kernel_eval(x_grid, theta, w, rho)
}
kernel_mean <- colMeans(kernel_draws_full)
kernel_lower <- apply(kernel_draws_full, 2, quantile, 0.05)
kernel_upper <- apply(kernel_draws_full, 2, quantile, 0.95)
plot(x_grid, kernel_mean, type = "l", lwd = 2,
     ylim = range(kernel_lower, kernel_upper),
     xlab = "x", ylab = "mu(x)",
     main = "Posterior kernel")

lines(x_grid, kernel_lower, lty = 2)
lines(x_grid, kernel_upper, lty = 2)

n_plot <- 200  # subset for plotting
draw_idx <- sample(1:n_draws, n_plot)

kernel_draws <- matrix(0, nrow = n_plot, ncol = length(x_grid))

for (i in seq_along(draw_idx)) {
  s <- samples[draw_idx[i], ]
  
  theta <- as.numeric(s[theta_cols])
  v <- as.numeric(s[v_cols])
  rho <- s["rho"]
  
  w <- compute_w(v, theta)
  
  kernel_draws[i, ] <- kernel_eval(x_grid, theta, w, rho)
}

matplot(x_grid, t(kernel_draws), type = "l", lty = 1,
        col = rgb(0,0,0,0.1),
        xlab = "x", ylab = "mu(x)",
        main = "Posterior kernel draws")
lines(x_grid, example_decay(x_grid), lwd = 3, col = "red")

```

