---
title: "Gamma Processes"
author: "Devon Grossett"
format: 
  html:
    css: styles.css
    theme: cosmo
    code-fold: false
bibliography: Hawkes_Bibtex.bib
---

## Dirichlet Process

::: {#def-dirichlet_process .definition}
## Dirichlet Process -- DP [@Muller2015_BayesianNonparametrics]

Let $\alpha > 0$ and $G_0$ be a probability measure defined on $S$. A DP with 
parameters $(\alpha, G_0)$ is a random probability measure $G$ defined on $S$
which assigns probability $G(B)$ to every (measurable) set $B$ such that for 
each (measurable) finite partition $\{B_1,\ldots,B_k\}$ of $S$, the vector 
$(G(B_1),\ldots,G(B_k))$ has the joint distribution. 
$$
(G(B_1),\ldots,G(B_k)) \sim \mathrm{Dir}(\alpha G_0(B_1),\ldots,\alpha G_0(B_k))
$$
where $\mathrm{Dir}(\cdot)$ is the Dirichlet distribution.

:::

The process is usually denoted as $\mathrm{DP}(\alpha G_0)$, or 
$\mathrm{DP}(\alpha, G_0)$. The parameter $\alpha$ is called the precision or 
total mass parameter and $G_0$ is the centering measure.

**Stick Breaking Construction**  A useful constructive definition of a DP 
random probability measure is given by Sethurman (1994) and based on the 
discrete nature of the process $G(\cdot) = \sum_{k=1}^\infty \tilde{w}_k \delta_{\theta_k}(\cdot)$.
Formally, let $\tilde{w}_k = v_k \prod_{\ell<k}(1-v_{\ell})$ with 
$v_{\ell}\stackrel{iid}{\sim} \mathrm{Beta}(1, \alpha)$ and $\theta_k\stackrel{iid}{\sim} G_0$, 
where $\{v_k\}$ and $\{\theta_k\}$ are independent. Then 
$$
G(\cdot) = \sum_{k=1}^\infty \tilde{w}_k\delta_{\theta_k}(\cdot)
$$ {#eq-dp_stickbreak}
defines a $\mathrm{DP}(\alpha G_0)$ random measure. 

In reality, we can choose some $k=K$ at which to truncate the series where 
beyond it, the values of $\tilde{w}_k$ become negligible. At that point, we can
set $v_K=1$, i.e. no stick breaking beyond this point.

```{r dp_stickbreak_func}
dp_stickbreak <- function(alpha, g0 = rexp, k_trunc = 100, ...) {
  # draw atoms from G_0
  theta <- g0(k_trunc, ...)
  
  # i.i.d. beta draws of v_k, truncate and set final value to 1
  v <- c(rbeta(k_trunc - 1, 1, alpha), 1)
  # stick remaining before each break
  remaining <- cumprod(c(1, 1 - v[-k_trunc]))
  # weights, with sum to one constraint
  w <- v*remaining
  w[k_trunc] <- 1 - sum(w[-k_trunc])
  
  return(list(theta = theta, weights = w, k_trunc = k_trunc))
}
```

An example with the base probability measure $H_0 \sim N(0,1)$, to visualise 
what the draws of a DP look like with varying $\alpha$.

```{r plot_dp_draws}
set.seed(1234)
par(mfrow = c(2, 2), mar = c(4, 5, 3, 1), oma = c(0, 0, 2, 0))

alphas <- c(1, 5, 20, 100)
truncs <- c(100, 200, 500, 2000) 

for (j in 1:4) {
  a <- alphas[j]
  k <- truncs[j]
  
  plot(NULL, xlim = c(-4, 4), ylim = c(0, 1),
       xlab = expression(t), ylab = expression(G(group("[", list(0, t), "]"))),
       main = bquote(alpha == .(a)))
  
  # add the theoretical Normal CDF (pnorm) in red
  curve(pnorm(x), from = -4, to = 4, col = "red", lwd = 2, add = TRUE)
  
  # draw 50 DP realisations
  for (i in 1:50) {
    # Using rnorm for the base measure g0
    dp <- dp_stickbreak(alpha = a, g0 = rnorm, k_trunc = k)
    ord <- order(dp$theta)
    lines(dp$theta[ord], cumsum(dp$weights[ord]), 
          type = "s", col = rgb(0, 0, 0, 0.1))
  }
}

# Add an overall title to the 2x2 grid
mtext(expression(paste("50 DP Realizations with ", G[0], " ~ N(0,1)")), 
      outer = TRUE, cex = 1.2, line = 0)
```

The Dirichlet Process is conjugate with respect to the sampling model 
$$
y_i \mid G \stackrel{\rm iid}{\sim} G,\quad i=1,\ldots,n
$$
such that if $G \sim \mathrm{DP}(\alpha G_0)$, then
$$
G \mid y_1, \ldots, y_n \sim \mathrm{DP}\left(\alpha+n, 
  \frac{\alpha G_0 + \sum_{i=1}^n \delta_{y_i}}{\alpha + n}\right)
$$

```{r dp_posterior_draws}
# function to sample from the posterior based on a DP(alpha, g0) prior
dp_post_stickbreak <- function(alpha, g0 = rexp, k_trunc = 100, obs, ...) {
  # helper function for simulation from centering measure of posterior DP
  rgn <- function(n_samples = 1, obs, alpha0, g00 = rnorm, ...) {
    prob_new <- alpha0/(alpha0 + length(obs))
    samples <- replicate(n_samples, {
      if (runif(1) < prob_new) {
        g00(1, ...)
      } else {
        sample(obs, 1)
      }
    })
  }
  dp_stickbreak(alpha = alpha + length(obs), g0 = rgn, k_trunc = k_trunc, 
                alpha0 = alpha, g00 = g0, obs = obs)
}
```

Lets imagine we observe some data from a mixture of normals 
$Y_i \sim 0.5N_1 + 0.5N_2$, $i=1,\ldots,n$ with 
$\mu_1=-1,\mu_2=1,\sigma_1^2=\sigma_2^2=0.5$ and $n=100$.

```{r plot_dp_posterior_draws}
set.seed(12345)

p <- rbinom(100, 1, 0.5)
y <- p*rnorm(100, -1, 0.5) + (1 - p)*rnorm(100, 1, 0.5)
grid_x <- seq(-4, 4, length.out = 500)
par(mfrow = c(2, 2), mar = c(4, 5, 3, 1), oma = c(0, 0, 2, 0))

for (j in 1:4) {
  a <- alphas[j]
  k <- truncs[j]
  
  plot(NULL, xlim = c(-4, 4), ylim = c(0, 1),
       xlab = expression(t), ylab = expression(G(group("[", list(0, t), "]"))),
       main = bquote(alpha == .(a)))
  
  all_cdfs <- matrix(0, nrow = 500, ncol = 50)
  
  # draw 50 posterior DP realisations
  for (i in 1:50) {
    dp <- dp_post_stickbreak(alpha = a, g0 = rnorm, k_trunc = k, obs = y)
    ord <- order(dp$theta)
    
    f_emp <- stepfun(dp$theta[ord], c(0, cumsum(dp$weights[ord])))
    all_cdfs[, i] <- f_emp(grid_x)
    
    lines(dp$theta[ord], cumsum(dp$weights[ord]), 
          type = "s", col = rgb(0, 0, 0, 0.1))
    
    legend("topleft", 
       legend = c("Posterior Mean", "True G"), 
       col = c("black", "red"), 
       lwd = c(1, 2), 
       # lty = c(1, 2),
       bty = "n",  
       cex = 0.8) 
  }
  
  curve(0.5*pnorm(x, -1, 0.5) + 0.5*pnorm(x, 1, 0.5), 
        from = -4, to = 4, col = "red", lwd = 2, add = TRUE)
  
  mean_cdf <- rowMeans(all_cdfs)
  lines(grid_x, mean_cdf, col = "black")
}

mtext(expression(paste("50 Posterior DP Realizations from prior DP(", alpha, ", N(0,1))")), 
      outer = TRUE, cex = 1.2, line = 0)
```

## Gamma Process Prior

::: {#def-gamma_process .definition}
## Gamma Process [@Phadia2016_PriorProcesses]

Let $\{Z_t\}_{t\geq0}$ be a stochastic process such that (i) $Z_0=0$, (ii) $Z_t$
has independent increments in non-overlapping intervals; and (iii) for $t>s$, 
the increment 
$$
(Z_t - Z_s) \sim \mathrm{Gamma}(c(\alpha(t)-\alpha(s)),c)
$$ 
where $c>0$ is a constant and $\alpha(t)$, $t>0$ be an increasing left-continuous 
function such that $\alpha(0)=0$. Then $Z_t$ is said to be a Gamma Process with 
parameters $c\alpha(t)$ and $c$ and denoted $\mathrm{GaP}(c\alpha(t),c)$.
:::

Thus we can view a Gamma Process as an increasing process with random 
increments, where the amount of increase over any interval follows a Gamma 
distribution. In Bayesian applications, the random non-negative measure $G$ on
$\mathbb{R}^+$ created by the Gamma Process (where $G([0,t])=Z_t$) can be used 
as a prior over non-negative finite measures, written as
$$
G \sim \mathrm{GaPP}(\alpha H_0,\beta)
$$
where we have reparameterised to have shape parameter $\alpha>0$, rate parameter 
$\beta > 0$, and a base probability measure over the space $\Theta=\mathbb{R}^+$.

A draw from a Gamma Process can be constructed from a DP draw by first drawing a 
random probability measure $\tilde{G} \sim \mathrm{DP}(\alpha H_0)$, and then 
scaling the resulting weights by a random total mass 
$\gamma \sim \mathrm{Gamma}(\alpha,\beta)$.

Using the stick breaking construction of the DP, this leads to the following
representation:

$$
\begin{aligned}
\theta_k \mid H_0 &\stackrel{\rm iid}{\sim} H_0 \qquad k=1,2,\ldots\\
v_k \mid \alpha &\stackrel{\rm iid}{\sim} \mathrm{Be}(1,\alpha)\\
\tilde{w}_k &= v_k \prod_{\ell<k} (1-v_\ell)\\
\gamma &\sim \mathrm{Ga}(\alpha,\beta)\\
G({\rm d}u) &= \gamma\sum_{k=1}^\infty \tilde{w}_k \delta_{\theta_k}({\rm d}u)
\end{aligned}
$$ {#eq-gp_stickbreak}

Since the weights are stochastically decreasing in $k$, we can truncate our
draws of $v_k$ at some sufficiently large value of $K$, set the final weight 
$\tilde{w}_K$ to be one minus the sum of the previous $(1-K)$ weights,and 
represent the Gamma Process Prior as a finite sum with minimal impact on 
accuracy. This looks like:

$$
\begin{aligned}
\theta_k \mid H_0 &\stackrel{iid}{\sim} H_0 \qquad k=1,2,\ldots,K \\
v_k \mid \alpha &\stackrel{iid}{\sim} \mathrm{Be}(1,\alpha) \qquad k=1,2,\ldots,K-1 \\
\tilde{w}_k &= v_k \prod_{\ell < k} (1-v_\ell) \qquad k=1,2,\ldots,K-1 \\
\tilde{w}_K &= 1 - \sum_{k=1}^{K-1}\tilde{w}_k = \prod_{\ell=1}^{K-1}(1-v_\ell) \\
\gamma &\sim \mathrm{Ga}(\alpha,\beta) \\
G({\rm d}u) &= \gamma \sum_{k=1}^K \tilde{w}_k \delta_{\theta_k}({\rm d}u)
\end{aligned}
$$ {#eq-gp_stickbreak_truncate}

```{r gp_draws}
gp_from_dp <- function(alpha, beta, g0 = rexp, k_trunc = 100, ...) {
  gamma <- rgamma(1, alpha, rate = beta)
  dp <- dp_stickbreak(alpha = alpha, g0 = g0, k_trunc = k_trunc, ...)
  return(list(
    theta = dp$theta, 
    weights = gamma*dp$weights, 
    gamma = gamma,
    k_trunc = dp$k_trunc
  ))
}
```

An example of draws of a Gamma process with base measure 
$H_0 \sim \mathrm{Exp}(1)$, where we have shown plots of the cumulative measure
$G([0,t])$

```{r plot_gp_draws}
set.seed(123456)
par(mfrow = c(2, 2), mar = c(4, 5, 3, 1), oma = c(0, 0, 2, 0))

alphas <- c(1, 5, 20, 50)
betas <- alphas
truncs <- c(100, 200, 500, 2000) 

for (j in 1:4) {
  a <- alphas[j]
  b <- betas[j]
  k <- truncs[j]
  
  plot(NULL, xlim = c(0, 5), ylim = c(0, 2),
       xlab = expression(t), ylab = expression(G(group("[", list(0, t), "]"))),
       main = bquote(alpha == .(a) ~","~ beta ==.(b)))
  
  # add the theoretical Normal CDF (pnorm) in red
  curve(pexp(x), from = 0, to = 5, col = "red", lwd = 2, add = TRUE)
  
  # draw 50 DP realisations
  for (i in 1:50) {
    # Using rnorm for the base measure g0
    gp <- gp_from_dp(alpha = a, beta = b, g0 = rexp, k_trunc = k)
    ord <- order(gp$theta)
    lines(gp$theta[ord], cumsum(gp$weights[ord]), 
          type = "s", col = rgb(0, 0, 0, 0.1))
  }
}
```

## Hawkes Process with Dirichlet/Gamma Process Priors

### Univariate Hawkes Process

For a univariate Hawkes Process, the conditional intensity function is:
$$
\lambda^*(t) = \lambda_0 + \sum_{t_i<t}\mu(t-t_i)
$$

We will consider a GaPP for $\mu(\cdot)$ constructed so that any realisation of
it is a non-increasing function (a similar constraint to the common exponential 
and power decay Hawkes kernels). For convenience, we will parameterise the 
excitation function as 
$$
\mu(t) = \rho g(t)
$$
where $g(t)$ is a probability density function on $\mathbb{R}^+$. This 
parameterisation separates the amount of excitation, determined by the branching
ratio $\rho=\int_0^\infty \mu(t){\rm d}t$, from the shape of the excitation 
function with time, determined by $g(t)$. We will make the restriction 
$\rho \in (0,1)$ by choice of priors and proposals, to ensure a well behaved 
process (no exploding of events off to infinity).

To construct g(t), we start with a truncated Dirichlet Process
$$
\begin{aligned}
\theta_k \mid H_0 &\stackrel{iid}{\sim} H_0 \qquad k=1,2,\ldots,K \\
v_k \mid \alpha &\stackrel{iid}{\sim} \mathrm{Be}(1,\alpha) \qquad k=1,2,\ldots,K-1 \\
\tilde{w}_k &= v_k \prod_{\ell < k} (1-v_\ell) \qquad k=1,2,\ldots,K-1 \\
\tilde{w}_K &= 1 - \sum_{k=1}^{K-1}\tilde{w}_k = \prod_{\ell=1}^{K-1}(1-v_\ell) \\
G({\rm d}u) &=  \sum_{k=1}^K \tilde{w}_k \delta_{\theta_k}({\rm d}u)
\end{aligned}
$$ {#eq-dp_stickbreak_truncate}
We can construct a non-increasing function with the tail mass of the DP measure,
$\int_t^\infty G({\rm d}u) = \sum_{k=1}^K \tilde{w}_k I(\theta_k > t)$. Taking 
the integral over $(0,\infty)$:
$$
\begin{aligned}
\int_0^\infty \left( \sum_{k=1}^K \tilde{w}_k I(\theta_k > t) \right) {\rm d}t 
  &= \sum_{k=1}^K \tilde{w}_k \int_0^\infty I(t < \theta_k) {\rm d}t \\
  &= \sum_{k=1}^K \tilde{w}_k \theta_k
\end{aligned}
$$
Thereforce to create $g(t)$, we define new weights
$$
w_k = \frac{\tilde{w}_k}{\sum_j \tilde{w}_j \theta_j}
$$
and construct our normalised excitation function as
$$
g(t) = \sum_{k=1}^K w_k I(\theta_k > t)
$$
This has the desired normalisation condition, as well as being non-increasing in $t$.
Combined with the background intensity, $\lambda_0$, and the branching ratio, 
$\rho$, the Hawkes conditional intensity function is specified as
$$
\lambda^*(t \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v})) 
  = \lambda_0 + \rho \sum_{t_i<t} \sum_{k=1}^K w_k I(\theta_k > t - t_i)
$$ {#eq-hp_intensity_dpp}

We can swap the summation order in @eq-hp_intensity_dpp and note that 
$\sum_{t_i<t}I(t-t_i < \theta_k)$ is simply counting the number of events before
time $t$ that occured within the last $\theta_k$ time units. Therefore, this 
type of Prior for the Hawkes Process has the interpretation that each event that
occured within the last $\theta_k$ time units of a given time $t$ contributes
$\rho w_k$ to the conditional intensity, and this effect stacks for all 
atoms $\theta_k$ drawn from the Dirichlet Process Prior. Based on this
interpretation, we can define a counting process from that term of the 
conditional intensity
$$
N_k^*(t_i \mid \boldsymbol{\theta}) = \sum_{j=1}^{i-1}I(t_i-t_j < \theta_k) = \#\{j<i:t_i-t_j<\theta_k\}
$$

Before we calculate the compensator of a Hawkes Process of this form, we note
that:
$$
\begin{aligned}
\int_0^t \mu(u){\rm d}u &= \int_0^t \rho \sum_{k=1}^K w_k I(u < \theta_k) {\rm d}u \\
&= \rho \sum_{k=1}^K w_k \int_0^t I(u < \theta_k) {\rm d}u \\
&= \rho \sum_{k=1}^K \int_0^{\mathrm{min}\{t,\theta_k\}} {\rm d}u \\
&= \rho \sum_{k=1}^K w_k \mathrm{min}\{t,\theta_k\}
\end{aligned}
$$

Therefore, as the compensator of a Hawkes Process is 
$\Lambda^*(t) = \int_0^t \lambda^*(u) {\rm d}u = \lambda_0 t + \sum_{i=1}^{N_t}  M(t - t_i)$,
where $M(t) = \int_0^t\mu(u) {\rm d}u$, we can write the compensator of a 
Hawkes Process with this form of prior as: 
$$
\Lambda^*(T \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v})) 
  = \lambda_0 T + \rho \sum_{i=1}^{N_T}\sum_{k=1}^K w_k \min\{T-t_i,\theta_k\}
$$ {#eq-hp_compensator_gapp}
The likelihood for a Hawkes Process observed over the period $[0,T]$ with a
Gamma Process Prior for the excitation function is then
$$
\begin{aligned}
L &= \left[ \prod_{i=1}^{N_T} \lambda^*(t_i \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v})) \right] 
    e^{\Lambda^*(T \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v}))} \\
  &= \left[ \prod_{i=1}^{N_T} \left(\lambda_0 + \rho\sum_{k=1}^K w_k N_k^*(t_i \mid \boldsymbol{\theta})\right)\right] \\
  &\quad\times\exp\left\{\lambda_0 T +  \rho\sum_{k=1}^K w_k \sum_{i=1}^{N_T} \min\{T-t_i,\theta_k\}\right\} \\
\end{aligned}
$$ {#eq-hp_likelihood_gapp}
with log-likelihood
$$
\begin{aligned}
\ell
  &= \sum_{i=1}^{N_T} \log \lambda^*(t_i \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v}))
    - \Lambda^*(T \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v}))\\
  &= \sum_{i=1}^{N_T} \log \left(\lambda_0 + \rho\sum_{k=1}^K w_k N_k(t_i \mid \boldsymbol{\theta})\right) \\
  &\quad -\lambda_0 T + \rho\sum_{k=1}^K w_k \sum_{i=1}^{N_T} \min\{T-t_i,\theta_k\} \\
\end{aligned}
$$ {#eq-hp_loglikelihood_gapp}

### Inference

Our full model with chosen priors is

$$
\begin{aligned}
t_i \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{v} 
  &\stackrel{\rm iid}{\sim} \mathrm{HP}(\lambda^*(\cdot \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v})))
  \qquad i=1,\ldots,N_T \\
  \theta_k \mid H_0(\cdot \mid \phi) &\stackrel{\rm iid}{\sim} \mathrm{Exp}(\phi) \qquad k=1,\ldots,K \\
  v_k \mid \alpha &\stackrel{\rm iid}{\sim} \mathrm{Beta}(1,\alpha) \qquad k=1,\ldots,K-1 \\
  \lambda_0 \mid \nu &\sim \mathrm{Exp}(\nu) \\
  \rho \mid r_1, r_2 &\sim \mathrm{Beta}(r_1, r_2) \\
  \alpha \mid a_1, a_2 &\sim \mathrm{Ga}(a_1, a_2) \\
  \phi \mid f_1, f_2 &\sim \mathrm{Ga}(f_1, f_2) \\
\end{aligned}
$$
and has full conditional posterior distributions

#### $\lambda_0$:

$$
\begin{aligned}
p(\lambda_0 \mid -) 
  &\propto p(\mathbf{t} \mid -)p(\lambda_0 \mid \nu) \\
  &\propto \left[ \prod_{i=1}^{N_T} \lambda^*(t_i \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v})) \right]
    e^{-\Lambda^*(T \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v}))} \\
  &\qquad \times e^{-\nu \lambda_0} \\
  &\propto \left[ \prod_{i=1}^{N_T} \lambda^*(t_i \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v})) \right] 
    e^{-\lambda_0(T + \nu)}
\end{aligned}
$$

We will use M-H updates of $\tilde{\lambda}_0 = \log\lambda_0$ with Normal proposal: 
$\mathrm{N}(\tilde{\lambda}_0,\sigma_{\lambda_0}^2)$. This transformation has Jacobian 
$\left|\frac{{\rm d}\lambda_0}{{\rm}d\tilde{\lambda}_0}\right| = \lambda_0$, and therefore
$$
\begin{aligned}
p(\tilde{\lambda}_0 \mid -) 
  &= p(\lambda_0 = e^{\tilde{\lambda}_0} \mid -)\left|\frac{{\rm d}\lambda_0}{{\rm}d\tilde{\lambda}_0}\right| \\
  &= p(\lambda_0 = e^{\tilde{\lambda_0}} \mid -) \lambda_0 \\
\log p(\tilde{\lambda}_0 \mid -) &= \log p(\lambda_0 = e^{\tilde{\lambda}_0} \mid -) + \log \lambda_0
\end{aligned}
$$

As the proposal distribution is symmetric about $\tilde{\lambda_0}$, the 
acceptance ratio is simply 
$$
r = \frac{p(\tilde{\lambda}_0^* \mid -)}{p(\tilde{\lambda}_0 \mid -)}
$$

#### $\rho$:

$$
\begin{aligned}
p(\rho \mid -) 
  &\propto p(\mathbf{t} \mid -)p(\rho \mid r_1, r_2) \\
  &\propto \left[ \prod_{i=1}^{N_T} \lambda^*(t_i \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v})) \right]
    e^{-\Lambda^*(T \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v}))} \\
  &\qquad \times \rho^{r_1 - 1} (1-\rho)^{r_2-1}
\end{aligned}
$$

We will use M-H updates of $\tilde{\rho} = \text{logit}(\rho)$ with Normal proposal: 
$\mathrm{N}(\tilde{\rho},\sigma_\rho^2)$. This transformation has Jacobian 
$\left|\frac{{\rm d}\rho}{{\rm}d\tilde{\rho}}\right| = \rho(1-\rho)$, and therefore
$$
\begin{aligned}
p(\tilde{\rho} \mid -) 
  &= p(\rho = \text{logit}^{-1}(\tilde{\rho}) \mid -)\left|\frac{{\rm d}\rho}{{\rm}d\tilde{\rho}}\right| \\
  &= p(\rho = \text{logit}^{-1}(\tilde{\rho}) \mid -) \rho(1-\rho) \\
\log p(\tilde{\rho} \mid -) &= \log p(\rho = \text{logit}^{-1}(\tilde{\rho}) \mid -) + \log\rho + \log(1-\rho)
\end{aligned}
$$

As the proposal distribution is symmetric, the acceptance ratio is simply 
$$
r = \frac{p(\tilde{\rho}^* \mid -)}{p(\tilde{\rho} \mid -)}
$$

#### $\theta_k$:

$$
\begin{aligned}
p(\theta_k \mid -) 
  &\propto p(\mathbf{t} \mid -) p(\theta_k \mid \phi) \\
  &\propto \left[ \prod_{i=1}^{N_T} \lambda^*(t_i \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v})) \right]
    e^{-\Lambda^*(T \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v}))} \\
  &\qquad \times e^{-\phi\theta_k} \\
\end{aligned}
$$

The sampling scheme for this comes from @Arnold2020_BayesNonparametricBathtub, 
where mixing is assisted by selecting an atom at random according to the size of
the unscaled weights $\tilde{w}$. For each pass of the sampler we update 
$k_\text{sim} \leq K$ randomly selected atoms (selected with replacement). This 
ensures that the most influential atoms are updated with greatest frequency.

Of the selected atoms, we update $\tilde{\theta}_k = \log\theta_k$ with Normal
proposal: $\mathrm{N}(\tilde{\theta}_k,\sigma_\theta^2)$.

As the proposal distribution is symmetric, the acceptance ratio is simply 
$$
r = \frac{p(\tilde{\theta}_k^* \mid -)}{p(\tilde{\theta}_k \mid -)}
$$

#### $v_k$:

$$
\begin{aligned}
p(v_k \mid -) 
  &\propto p(\mathbf{t} \mid -) p(v_k \mid \alpha) \\
  &\propto \left[ \prod_{i=1}^{N_T} \lambda^*(t_i \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v})) \right]
    e^{-\Lambda^*(T \mid \lambda_0, \rho, \boldsymbol{\theta}, \mathbf{w}(\mathbf{v}))} \\
  &\qquad \times (1-v_k)^{\alpha-1} \\
\end{aligned}
$$

As we did for $\rho$, the M-H update will be of
$\tilde{v}_k = \text{logit}(v_k)$ with Normal proposal: 
$\mathrm{N}(\tilde{v}_k,\sigma_v^2)$. 

As the proposal distribution is symmetric, the acceptance ratio is simply 
$$
r = \frac{p(\tilde{v}_k^* \mid -)}{p(\tilde{v}_k \mid -)}
$$

#### $\alpha$:

$$
\begin{aligned}
p(\alpha \mid -) 
  &\propto p(\mathbf{v} \mid \alpha)p(\alpha) \\
  &\propto \left[ \prod_{k=1}^{K-1} \alpha (1-v_k)^{\alpha-1} \right] \alpha^{a_1 - 1}e^{-a_2\alpha} \\
  &\propto \alpha^{a_1 + K - 2} \tilde{w}_K^\alpha e^{-a_2\alpha} \\
  &\propto \alpha^{(a_1 + K - 1) - 1} e^{-\alpha(a_2 - \log\tilde{w}_K)} \\
\end{aligned}
$$

Therefore the Gibbs update is 
$$
\alpha \mid - 
  \sim \mathrm{Ga}\left(a_1+K-1, a_2-\log\tilde{w}_K\right)
$$

#### $\phi$:

$$
\begin{aligned}
p(\phi \mid -) 
  &\propto p(\boldsymbol{\theta} \mid \phi)p(\phi \mid f_1 f_2) \\
  &\propto \left[\prod_{k=1}^K \phi e^{-\phi \theta_k}\right]\phi^{f_1-1}e^{-\phi f_2} \\
  &\propto \phi^K e^{-\phi\sum_{k=1}^K\theta_k}\phi^{f_1-1}e^{-\phi f_2} \\
  &\propto \phi^{(f_1 + K)-1}e^{-\phi\left(f_2 + \sum_{k=1}^K\theta_k\right)}
\end{aligned}
$$

Therefore the Gibbs update is 
$$
\phi \mid - 
  \sim \mathrm{Ga}\left( f_1 + K, f_2 + \sum_{k=1}^K\theta_k \right)
$$

### Demonstration

I have put all the code for posterior functions, Gibbs and MH updates, and a 
parent function that coordinates sampling for a single MCMC chain into 
`./R/univariate_hawkes_npb.R` 

```{r}
source("./R/univariate_hawkes_npb.R", echo = TRUE)

# project library, contains the `sim_hp` function used below
library(HawkesProject)

# generate a random hawkes kernel from the DP construction above
set.seed(7) # reproducibility
dp <- dp_stickbreak(alpha = 2, g0 = rexp, k_trunc=40, rate = 2)
example_decay <- function(x) {
  rho <- 0.6
  thetas <- dp$theta
  weights <- dp$weights/sum(dp$theta*dp$weights)
  vapply(x, function(xi) rho*sum(weights[thetas > xi]), numeric(1))
}

# visualise generated hawkes kernel
x_grid <- seq(0, 2.5, length.out = 500)
y <- example_decay(x_grid)
plot(x_grid, y, type = "s",
     xlab = "x", ylab = "Decay",
     main = "Example Decay Function")

# simulate process using generated hawkes kernel
set.seed(456)
simulated_process <- sim_hp(T_max = 200, lambda = 0.5, mu = example_decay)
plot(simulated_process)

# defining the priors and proposal distributions
prior_params <- list(a_1 = 1, a_2 = 1, f_1 = 1, f_2 = 1, r_1 = 4, r_2 = 4, nu = 0.1)
proposal_sds <- list(lambda0 = 0.2, rho = 1, theta_k = 2, v_k = 2)

# run MCMC chain (will do parallel chains later)
n_iter <- 2000
set.seed(6789)
init_list <- list(lambda0 = 1, rho = 0.5, theta = runif(20, 0, 5), 
                  v = runif(19, 0, 1), alpha = 1, phi = 2)
chains <- run_sampler(simulated_process$events, simulated_process$T_max, 
                      n_iter, init_list, prior_params, proposal_sds, 
                      progress = FALSE)
```

Visualising the posterior draws of the excitation function

```{r}
stickbreak <- function(v) {
  K <- length(v) + 1
  
  remaining <- cumprod(c(1, 1 - v))
  w_raw <- c(v, 1) * remaining
  
  # enforce sum to 1 numerically
  w_raw[K] <- 1 - sum(w_raw[-K])
  
  return(w_raw)
}

compute_w <- function(v, theta) {
  w_raw <- stickbreak(v)
  w <- w_raw / sum(w_raw * theta)
  return(w)
}

kernel_eval <- function(x, theta, w, rho) {
  sapply(x, function(xi) {
    rho * sum(w[theta > xi])
  })
}

samples <- chains$samples[-(1:100), ]
K <- 20
theta_cols <- paste0("theta", 1:K)
v_cols <- paste0("v", 1:(K-1))

x_grid <- seq(0, 5, length.out = 200)

n_draws <- nrow(samples)
kernel_draws_full <- matrix(0, nrow = n_draws, ncol = length(x_grid))
for (i in 1:n_draws) {
  s <- samples[i, ]
  
  theta <- as.numeric(s[theta_cols])
  v <- as.numeric(s[v_cols])
  rho <- s["rho"]
  
  w <- compute_w(v, theta)
  
  kernel_draws_full[i, ] <- kernel_eval(x_grid, theta, w, rho)
}
kernel_mean <- colMeans(kernel_draws_full)
kernel_lower <- apply(kernel_draws_full, 2, quantile, 0.05)
kernel_upper <- apply(kernel_draws_full, 2, quantile, 0.95)
plot(x_grid, kernel_mean, type = "l", lwd = 2,
     ylim = range(kernel_lower, kernel_upper),
     xlab = "x", ylab = "mu(x)",
     main = "Posterior kernel")

lines(x_grid, kernel_lower, lty = 2)
lines(x_grid, kernel_upper, lty = 2)

n_plot <- 200  # subset for plotting
draw_idx <- sample(1:n_draws, n_plot)

kernel_draws <- matrix(0, nrow = n_plot, ncol = length(x_grid))

for (i in seq_along(draw_idx)) {
  s <- samples[draw_idx[i], ]
  
  theta <- as.numeric(s[theta_cols])
  v <- as.numeric(s[v_cols])
  rho <- s["rho"]
  
  w <- compute_w(v, theta)
  
  kernel_draws[i, ] <- kernel_eval(x_grid, theta, w, rho)
}

matplot(x_grid, t(kernel_draws), type = "l", lty = 1,
        col = rgb(0,0,0,0.1),
        xlab = "x", ylab = "mu(x)",
        main = "Posterior kernel draws")
lines(x_grid, example_decay(x_grid), lwd = 3, col = "red")

```

