---
title: "Hawkes Notes & Exploration"
author: "Devon Grossett"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    css: styles.css
    theme: cosmo
    include-in-header:
      text: |
        <script>
        window.MathJax = {
          tex: {
            macros: {
              dd: "{\\mathrm{d}}"
            }
          }
        };
        </script>
        <script id="MathJax-script" async src="https://cdn.jsdelivr.net"></script>
execute:
  freeze: auto
bibliography: Hawkes_Bibtex.bib
crossref:
  eq-prefix: "Equation"
  thm-prefix: "Theorem"
  def-prefix: "Definition"
---

These are some notes taken while reading through @Laub2021_ElementsHawkes

```{r}
# WIP library that currently has my simulation and plotting functions
library(HawkesProject)
```

```{r, include=FALSE}
# --- Color Palettes ---

# NORD (Minimalist / Clean)
NORD_BLUE   <- "#81A1C1" # Primary Intensity
NORD_SLATE  <- "#4C566A" # Baseline / N(t)
NORD_RED    <- "#BF616A" # Events / Ticks
NORD_FROST  <- "#88C0D0" # Secondary Line
NORD_NIGHT  <- "#2E3440" # Titles / Text
```



## Point Process Background

### Counting and Point Process Background

Some definitions that are used throughout

::: {#def-counting_process .definition}
## Counting Process and History

A *counting process* is a stochastic process $(N(t): t \geq 0)$ (often shortened 
to $N_t$) taking values in $\mathbb{N}_0$ that satisfy $N(0) = 0$, is a.s. 
finite, and is a right continuous step function with increments of size $+1$

Further, denote by $(\mathcal{H}(u): u \geq 0)$ the *history* of the arrivals up
to time $u$. $\mathcal{H}(u)$ is a filtration, that is, an increasing sequence 
of $\sigma$-algebras.
:::

We can also consider the arrival times $T = \{ T_1, T_2, \ldots \}$ at which the 
process $N(t)$ jumps. The process defined as these arrival times is called a 
point process.

::: {#def-point_process .definition}
## Point Process

If a sequence of random variables $\mathbf{T} = \{T_1, T_2, \ldots \}$, taking 
values in $[0, \infty)$, has $\mathbb{P}(0 \leq T_1 \leq T_2 \leq \ldots) = 1$, 
and the number of points in a bounded region is a.s. finite, then $\mathbf{T}$ 
is a (*simple*) point process.
:::

These terms are often used interchangeably, although it should be clear from the
context whether $N(t)$ or $\mathbf{T}$ is being discussed.

::: {#def-poisson_process .definition}
## Poisson Process

A counting process $(N(t): t \geq 0)$ is a homogeneous Poisson process with rate
$\lambda > 0$ if

1.  For any interval $I$, $N(I) \sim \mathrm{Pois}(\lambda |I|)$
2.  For any $n$ disjoint intervals $I_1, I_2, \ldots, I_n$, the random variables
$N(I_1), N(I_2), \ldots, N(I_n)$ are independent.
:::

From this definition we get $$
\mathbb{P}(N(t) = n) = \frac{(\lambda t)^n e^{-\lambda t}}{n!}, \quad 
n=0,1,2,\ldots
$$ {#eq-pois_dist}

### Conditional Intensity Functions

Given the history up until the last arrival $u$, $\mathcal{H}(u)$, define the 
conditional cumulative distribution (CDF) function of the next arrival time 
$T_{k+1}$ as 
$$
F(t \mid \mathcal{H}(u)) = \int_u^t \mathbb{P}(T_{k+1} \in [s, s + \dd s] 
\mid \mathcal{H}(u)) \, \dd s = \int_u^t f(s \mid \mathcal{H}(u)) \, \dd s
$$

The joint PDF for realisations $\{ t_1, t_2, \ldots, t_k \}$ is then, by the 
chain rule for probabilities $$
f(t_1, t_2, \ldots, t_k) = \prod_{i=1}^k f(t_i \mid \mathcal{H}(t_{i-1}))
$$ {#eq-joint_pdf}

It's very common to suppress notation of conditioning upon $\mathcal{H}(u)$ 
using a superscript asterisk notation, e.g. 
$f^*(t) \coloneqq f(t \mid \mathcal{H}(u))$.

It is often more convenient to work with the conditional intensity function 
(also known as the hazard function in some fields) $\lambda^*(t)$ defined as 
$$
\lambda^*(t) = \frac{f^*(t)}{1 - F^*(t)}
$$ {#eq-hazard}

However there is another representation for the conditional intensity as the 
expected rate of arrivals conditioned on $\mathcal{H}(t)$:

::: {#def-conditional_intensity .definition}
## Conditional Intensity Function

Consider a counting process $N$ with associated histories $\mathcal{H}$. If a 
non-negative function $\lambda^*(t)$ exists such that 
$$
\lambda^*(t) = \lim_{h \downarrow 0} \frac{\mathbb{E}[N(t + h) - N(t) 
\mid \mathcal{H}(t)]}{h}
$$ which only relies of information of $N$ in the past (that is, 
$\lambda^*(t)$ is $\mathcal{H}(t)$-measurable), then it is called the 
conditional intensity function of $N$.
:::

Lastly, we define the compensator, which is frequently used in parameter
estimation and goodness of fit testing of point processes.

::: {#def-compensator .definition}
## Compensator

For a counting process $N$ the non-decreasing function 
$$
\Lambda(t) = \int_0^t \lambda^*(s) \dd s
$$ is called the compensator of the counting process.
:::

## Hawkes Process Background

The Hawkes process is a model for a ‘self-exciting’ counting process, where the 
history of past arrivals increases the chance of subsequent arrivals for a
period of time.

The Hawkes process is defined by its conditional intensity function (originally
given in @Hawkes1971_SelfExcitingSpectra)

::: {#def-hawkes_process .definition}
## Hawkes Process

A Hawkes process is a counting process $(N(t): t \geq 0)$ whose conditional 
intensity function for $t \ge 0$ is 
$$
\lambda^{*}_{t} = \lambda + \sum_{t_i<t} \mu (t - t_i)
$$ {#eq-hp_lambda} 
where $\lambda > 0$ is the background arrival rate, 
$\mu : (0,\infty) \rightarrow [0,\infty)$ is the excitation function (or 
kernel), and $t_i$ are all the arrival times in the history of the process 
$\mathcal{H}(u)$.
:::

The most common choice of $\mu$ is the exponentially decaying kernel 
$$
\mu(t) = \alpha e^{-\beta t} , \quad \alpha,\beta > 0
$$ {#eq-hp_exp_kernel}

An alternative way to represent a Hawkes process is through a branching process
known as an immigrant-birth process (@Hawkes1974_ClusterRepresentation). In this
view, the immigration process is a homogeneous Poisson process with rate 
$\lambda$, and the arrival of an individual from the immigrant process at time 
$s$ initiates a non-homogeneous Poisson process of births with intensity 
$\mu(t - s)$ for $t > s$, which subsequently may result in further descendants. 
Each arrival generates a Poisson-distributed number of first-generation 
offspring with mean $\eta$, where $\eta$ is known as the branching ratio and is 
given by

$$
\eta = \int_0^{\infty}\mu(t)dt
$$ {#eq-hp_branchingratio}

For the process to be stationary with finite mean, the condition $\eta < 1$ is 
necessary and sufficient.

## Maximum Likelihood Estimation

#### Example: Exponential MLE  {.unnumbered}

Below is an example of estimating a parameter from $\mathrm{Exp(\lambda)}$ data 
using Maximum Likelihood Estimation (MLE). For this problem
$$
\begin{aligned}
L(\lambda \mid \mathbf{t}) &= \prod_{i=1}^n \lambda e^{-\lambda t_i} \\
l(\lambda \mid \mathbf{t}) &= n \log \lambda - \lambda \sum_{i=1}^n t_i
\end{aligned}
$$

```{r exp_mle}
# generate Exp(λ) variables, with each column a new realisation
generate_exp_matrix <- function(n_obs, n_trials, lambda_true, seed = NULL) {
  # set seed if given
  if (!is.null(seed)) {set.seed(seed)}
  
  # simulate matrix of exponential samples
  matrix(
    rexp(n_obs*n_trials, rate = lambda_true),
    nrow = n_obs,
    ncol = n_trials
  )
}

# log-likelihood of Exp(λ) data
exp_loglik <- function(lambda, t) {
  n <- length(t)
  n*log(lambda) - lambda*sum(t)
}

# since optim minimises by default
exp_negloglik <- function(lambda, t) {
  -exp_loglik(lambda, t)
}

# take t matrix and create a dataframe of mle parameter estimates and std errors
mle_results <- function(t_mat) {
  n <- ncol(t_mat)
  
  # to store results
  results <- data.frame(
    trial = 1:n, 
    sample_size = rep(nrow(t_mat), n),
    mean = rep(NA, n),
    se = rep(NA, n),
    lower = rep(NA, n),
    upper = rep(NA, n)
  )
  
  # mle fit for each trial
  for (i in 1:n) {
    fit <- optim(
      par = 1,
      fn = exp_negloglik,
      t = t_mat[, i],
      method = "L-BFGS-B",
      lower = 1e-8,
      hessian = TRUE
    )
    results[i, "mean"] <- fit$par
    results[i, "se"] <- sqrt(1 / fit$hessian) 
  }
  results$lower <- results$mean - 1.96*results$se
  results$upper <- results$mean + 1.96*results$se
  
  results
}

lambda_true <- 2
t_mat <- generate_exp_matrix(
  n_obs = 100, 
  n_trials = 500, 
  lambda_true = lambda_true, 
  seed = 5
)
results <- mle_results(t_mat)
```

```{r exp_mle_results_table, echo=FALSE}
summary_table <- data.frame(
  Parameter = "λ",
  MC_Trials = nrow(results),
  Sample_Size = max(results$sample_size),
  Mean = mean(results$mean),
  Bias = mean(results$mean) - lambda_true,
  # SD = sd(results$mean),
  SD = sqrt(mean((results$mean - mean(results$mean))^2)),
  RMSE = sqrt(mean((results$mean - lambda_true)^2)),
  Coverage_95 = mean(results$lower <= lambda_true &
                  lambda_true <= results$upper),
  MC_Error = sd(results$mean) / sqrt(nrow(results))
)

knitr::kable(
  summary_table,
  digits = c(NA, 0, 0, 4, 4, 4, 4, 4, 4),
  col.names = c(
    "Parameter", 
    "MC Trials",
    "Sample Size",
    "Mean", 
    "Bias", 
    "SD", 
    "RMSE", 
    "95% Coverage", 
    "MC Error"
  )
)
```


```{r exp_mle_plot, echo=FALSE}
cover <- results$lower <= 2 & results$upper >= 2
cols <- ifelse(cover, NORD_SLATE, NORD_RED)
widths <- ifelse(cover, 0.5, 1.5) 

plot_base_canvas(
  xlim = c(min(results$trial), max(results$trial)), 
  ylim = range(results$lower, results$upper),
  title = "Coverage of 95% CI",
  xlab = "Simulation",
  ylab = expression(hat(lambda))
)
points(results$trial, results$mean, col = cols, pch = 16, cex = 0.6)
segments(
  results$trial, results$lower,
  results$trial, results$upper,
  col = cols, lwd = widths
)

# plot(
#   results$trial,
#   results$mean,
#   pch = 16,
#   cex = 0.6,
#   col = cols,
#   ylim = range(results$lower, results$upper),
#   xlab = "Simulation",
#   ylab = expression(hat(lambda)),
#   main = "Coverage of 95% CI"
# )

# segments(
#   results$trial, results$lower,
#   results$trial, results$upper,
#   col = cols, lwd = widths
# )

abline(h = 2, lwd = 2)
```

::: {#thm-point_process_likelihood}
## Point Process Likelihood

Let $N$ be a simple point process with conditional intensity $\lambda^*(t)$ and
compensator $\Lambda(t)$. If we observe all the arrival times over the time 
period $[0, T ]$, denoted $\{t_1, \ldots, t_{N(T)} \}$, then the likelihood 
function $L(\mathbf{\theta} \mid \mathbf{t})$ for $N$ is 
$$
L(\mathbf{\theta} \mid \mathbf{t})= \left[ \prod_{i=1}^{N_T} \lambda^*(t_i 
  \mid \mathbf{\theta}, \mathbf{t}) \right] 
  e^{-\Lambda(T \mid \mathbf{\theta}, \mathbf{t})}
$$ {#eq-pp_likelihood}
and the log-likelihood function is
$$
l(\mathbf{\theta} \mid \mathbf{t}) = \sum_{i=1}^{N_T} \log(\lambda^*(t_i \mid
  \mathbf{\theta}, \mathbf{t})) - \Lambda(T \mid \mathbf{\theta}, \mathbf{t})
$$ {#eq-pp_loglikelihood}
:::

For brevity we will not keep writing the observation vector $\mathbf{t}$ and the
parameter vector $\mathbf{\theta}$, so in that notation 
$L = L(\mathbf{\theta} \mid \mathbf{t})$, 
$l = l(\mathbf{\theta} \mid \mathbf{t})$, 
$\lambda^*(t_i) = \lambda^*(t_i \mid \mathbf{\theta}, \mathbf{t})$, and
$\Lambda(t) = \Lambda(t \mid \mathbf{\theta}, \mathbf{t})$.


#### Example: Poisson Process MLE  {.unnumbered}

For a simple example we'll estimate a Poisson process' parameter using MLE and
the form of the log-likelihood in @eq-pp_loglikelihood. In this case, 
$\lambda^*(t) = \lambda$ and the compensator $\Lambda(t) = \lambda t$. Therefore
$$
l = N_T \log \lambda - \lambda T
$$

The value of $\lambda$ that maximises this is easily found by setting 
$\frac{\dd l}{\dd \lambda}=0$ and solving for $\lambda$, giving 
$$
\hat{\lambda} = N_T/T
$$

```{r poissonprocess_mle}
lambda <- 2
T_max <- 10
set.seed(6)

# simulate a Poisson process over [0, T_max] using custom library
arrival_times <- sim_pp(T_max, lambda)
plot(arrival_times)

# log-likelihood of Poisson process arrival time data
poisprocess_loglik <- function(lambda, arrival_times, T_max) {
  N_T <- length(arrival_times)
  N_T*log(lambda) - lambda*T_max
}

# since optim minimises by default
poisprocess_negloglik <- function(lambda, arrival_times, T_max) {
  -poisprocess_loglik(lambda, arrival_times, T_max)
}

fit <- optim(
  par = 1, 
  fn = poisprocess_negloglik, 
  arrival_times = arrival_times$events,
  T_max = T_max,
  method = "L-BFGS-B",
  lower = 1e-8,
  hessian = TRUE
)

cat(sprintf(
  "Analytic Solution = %.3f ± %.3f \nNumeric MLE = %.3f ± %.3f", 
  length(arrival_times)/T_max, 1.96*sqrt(length(arrival_times))/T_max, 
  fit$par, 1.96/sqrt(fit$hessian)
))
```



### Hawkes Process MLE
To get the form of the Hawkes likelihood, first note that the integral in 
$\Lambda(T)$ over $[0, T]$ can be broken down into 
$$
[0, t_1] \cup (t_1,t_2] \cup \ldots \cup (t_{N_T - 1}, t_{N_T}] 
  \cup (t_{N_T}, T]
$$ 
and therefore 
$$
\Lambda(T) = \int_0^{t_i} \lambda^*(s) \dd s + \sum_{i=1}^{N_T-1} 
  \int_{t_i}^{t_i + 1} \lambda^*(s) \dd s + \int_{t_{N_T}}^{T} \lambda^*(s) \dd s
$$

Substituting the Hawkes conditional intensity in @eq-hp_lambda, and using 
$M(t) \coloneqq \int_0^t \mu(s) \dd s$, the integral term can be written as 
$$
\begin{aligned}
\int_{t_i}^{t_{i+1}} \lambda^*(s) \dd s &= \int_{t_i}^{t_{i+1}} 
  \left[ \lambda + \sum_{t_j < s} \mu(s - t_j) \right] \dd s 
  \\
&= \lambda (t_{i+1} - t_i) + \sum_{j=1}^i \int_{t_i}^{t_{i+1}} \mu(s - t_j) 
  \dd s 
  \\
&= \lambda (t_{i+1} - t_i) + \sum_{j=1}^i \left[ M(t_{i+1} - t_j) - 
  M(t_i - t_j) \right]
\end{aligned}
$$

This summation holds for $i=1,\ldots,N_T-1$, for $i=0$ where the summation term
disappears due to no events in the history, and for $i = N_T$ if we abuse 
notation a little and say $T=t_{N_T+1}$. Therefore we can write
$$
\begin{aligned}
\Lambda(T) &= \sum_{i=1}^{N_T} \left\{ \lambda(t_{i+1} - t_i) + 
  \sum_{j=1}^i [M(t_{i+1} - t_j) - M(t_i - t_j)] \right\} 
  \\ 
&= \lambda T + \sum_{i=1}^{N_T} \sum_{j=1}^i [M(t_{i+1} - t_j) - M(t_i - t_j)]
  \\
&= \lambda T + \sum_{i=1}^{N_T}  [M(T - t_i) - M(0)] 
  = \lambda T + \sum_{i=1}^{N_T}  M(T - t_i)
\end{aligned}
$$ {#eq-hawkes_compensator}

The log-likelihood (@eq-pp_loglikelihood), combined with the this form of the 
Hawkes compensator $\Lambda(T)$ and the Hawkes conditional intensity 
(@eq-hp_lambda) gives us

$$
l = \sum_{i=1}^{N_T} \log \left[ \lambda + \sum_{j=1}^{i-1} \mu(t_i - t_j)  
  \right] - \lambda T - \sum_{i=1}^{N_T}  M(T - t_i)
$$
The double sum will mean that for a numeric optimisation routine, we will have 
to evaluate a $\mathcal{O}([N(T)]^2)$ function for each optimisation step which 
will scale poorly.

#### Example: Exponential Hawkes Process MLE  {.unnumbered}

For a Hawkes process with an exponential kernel (@eq-hp_exp_kernel), the 
log-likelihood function is
$$
l = \sum_{i=1}^{N_T} \ln \left[ \lambda + \alpha \sum_{j=1}^{i-1} 
  e^{-\beta (t_i - t_j)} \right] - \Lambda (T)
$${#eq-exp_hp_loglik}
where the compensator of the process is, using @eq-hawkes_compensator, 
$$
\Lambda(t) = \lambda T + \sum_{i=1}^{N_T}M(T-t_i) 
  = \lambda T + \frac{\alpha}{\beta} \sum_{i=1}^{N_T} 
  \left[ 1 - e^{-\beta (T - t_i)} \right]
$$

The inner sum in @eq-exp_hp_loglik can simplified for faster computation. In 
particular, for $i \in \left\{2,...,N_T \right\}$, let 
$A(i) = \sum_{j=1}^{i-1} e^{-\beta(t_i - t_j)}$ (the inner sum) so that we can 
find the recursive relation
$$
\begin{aligned}
A(i) &= \sum_{j=1}^{i-1} e^{-\beta(t_i - t_j)} = \sum_{j=1}^{i-1} 
  e^{-\beta(t_i - t_j + t_{i-1} - t_{i-1})} \\
&= e^{-\beta(t_i - t_{i-1})} \sum_{j=1}^{i-1} e^{-\beta(t_{i-1} - t_j)} 
  = e^{-\beta(t_i - t_{i-1})}\left(1 +  \sum_{i=1}^{j-2} 
  e^{-\beta(t_{i-1} - t_j)}  \right) \\
&= e^{-\beta(t_i - t_{i-1})}(1 + A(i-1))
\end{aligned}
$$

With the base case that $A(1) = 0$, the exponential Hawkes likelihood 
(@eq-exp_hp_loglik) can be rewritten
$$
l = \sum_{i=1}^{N_T} \ln \left[ \lambda + \alpha A(i) \right] - \Lambda (T)
$$ {#eq-exp_hp_loglik_fast}

```{r hawkesprocess_mle}

exp_hp_intensity <- function(theta, t, H_t) {
  # unpack parameters
  lambda <- theta[1]
  alpha <- theta[2]
  beta <- theta[3]
  
  lambda + alpha*sum(exp(-beta*(t - H_t)))
}

exp_hp_compensator <- function(theta, t, H_t) {
  # unpack parameters
  lambda <- theta[1]
  alpha <- theta[2]
  beta <- theta[3]
  
  if (t <= 0) return(0)
  lambda*t + (alpha/beta)*sum(1 - exp(-beta*(t - H_t)))
}

exp_hp_loglik <- function(theta, T_max, H_t) {
  # there is a speedup for the exponential hawkes log-likelihood that is O(N(T_max))
  # this is the more expensive O([N(T_max)]^2) computation
  log_intensities <- sapply(H_t, function(t_i) {
    past_events <- H_t[H_t < t_i] # Find history by comparing values
    log(exp_hp_intensity(theta = theta, t = t_i, H_t = past_events))
  })
  
  # total log-likelihood = sum(log(intensities)) - compensator at T
  sum(log_intensities) - exp_hp_compensator(theta = theta, t = T_max, H_t = H_t)
}

exp_hp_loglik_fast <- function(theta, T_max, H_t) {
  lambda <- theta[1]
  alpha <- theta[2]
  beta <- theta[3]
  
  n <- length(H_t)
  # Recursive computation of the summation term A(i)
  A <- numeric(n)
  if (n > 1) {
    for (i in 2:n) {
      A[i] <- exp(-beta * (H_t[i] - H_t[i-1])) * (1 + A[i-1])
    }
  }
  
  log_intensities <- log(lambda + alpha*A)
  sum(log_intensities) - exp_hp_compensator(theta, T_max, H_t)
}

exp_hp_negloglik<- function(theta, T_max, H_t, fast=TRUE) {
  if (fast) return(-exp_hp_loglik_fast(theta = theta, T_max = T_max, H_t = H_t))
  else return(-exp_hp_loglik(theta = theta, T_max = T_max, H_t = H_t))
}

T_max <- 200; lambda <- 0.5; alpha <- 2; beta <- 3;
set.seed(123)
t <- sim_hp(T_max = T_max, lambda = lambda, mu = function(t) {alpha*exp(-beta*t)})

fit <- optim(
  par = c(1, 1, 1), 
  fn = exp_hp_negloglik, 
  H_t = t$events,
  T_max = T_max,
  method = "L-BFGS-B",
  lower = c(1e-8, 1e-8, 1e-8),
  hessian = TRUE
)

mle <- fit$par
v_cov <- solve(fit$hessian)
se <- sqrt(diag(v_cov))

cat("Maximum Likelihood Estimates with 95% CIs:\n", sprintf(
  "lambda_mle = %.3f ± %.3f \n alpha_mle = %.3f ± %.3f \n beta_mle = %.3f ± %.3f", 
  mle[1], 1.96*se[1], 
  mle[2], 1.96*se[2], 
  mle[3], 1.96*se[3]
))

```

```{r}
t_grid <- seq(0, T, length.out = 2000) 
intensities <- sapply(t_grid, function(t_i) {
  past <- t[t < t_i]
  lambda + sum(alpha * exp(-beta * (t_i - past)))
})

plot_base_canvas(c(0, T), c(0, max(intensities)*1.1), "Hawkes MLE Fit")
abline(h = fit$par[1], col = NORD_NIGHT, lty = 3) # Baseline
add_intensity(t_grid, intensities)
add_events(t)


```


## References {.unnumbered}
