---
title: "Hawkes Notes & Exploration"
author: "Devon Grossett"

format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    css: styles.css
    theme: cosmo

  pdf:
    documentclass: amsart
    number-sections: true
    keep-tex: true
    pdf-engine: xelatex
    header-includes:
      - \usepackage{amsmath}
      - \usepackage{amssymb}
      - \usepackage{mathtools}

execute:
  freeze: auto

bibliography: Hawkes_Bibtex.bib

crossref:
  eq-prefix: "Equation"
  thm-prefix: "Theorem"
  def-prefix: "Definition"
---

These are some notes taken while reading through @Laub2021_ElementsHawkes

```{r load_proj_lib, include=FALSE}
# WIP library that currently has my simulation and plotting functions
library(HawkesProject)
library(ggplot2)
library(sf)
library(rnaturalearth)
```

```{r colours, include=FALSE}
# --- Colour Palettes ---

# NORD (Minimalist / Clean)
NORD_BLUE   <- "#81A1C1" # Primary Intensity
NORD_SLATE  <- "#4C566A" # Baseline / N(t)
NORD_RED    <- "#BF616A" # Events / Ticks
NORD_FROST  <- "#88C0D0" # Secondary Line
NORD_NIGHT  <- "#2E3440" # Titles / Text
```



## Point Process Background

### Counting and Point Process Background

Some definitions that are used throughout

::: {#def-counting_process .definition}
## Counting Process and History

A *counting process* is a stochastic process $(N(t): t \geq 0)$ (often shortened 
to $N_t$) taking values in $\mathbb{N}_0$ that satisfy $N(0) = 0$, is a.s. 
finite, and is a right continuous step function with increments of size $+1$

Further, denote by $(\mathcal{H}(u): u \geq 0)$ the *history* of the arrivals up
to time $u$. $\mathcal{H}(u)$ is a filtration, that is, an increasing sequence 
of $\sigma$-algebras.
:::

We can also consider the arrival times $T = \{ T_1, T_2, \ldots \}$ at which the 
process $N(t)$ jumps. The process defined as these arrival times is called a 
point process.

::: {#def-point_process .definition}
## Point Process

If a sequence of random variables $\mathbf{T} = \{T_1, T_2, \ldots \}$, taking 
values in $[0, \infty)$, has $\mathbb{P}(0 \leq T_1 \leq T_2 \leq \ldots) = 1$, 
and the number of points in a bounded region is a.s. finite, then $\mathbf{T}$ 
is a (*simple*) point process.
:::

These terms are often used interchangeably, although it should be clear from the
context whether $N(t)$ or $\mathbf{T}$ is being discussed.

::: {#def-hpp .definition}
## Poisson Process

A counting process $(N(t): t \geq 0)$ is a homogeneous Poisson process with rate
$\lambda > 0$ if

1.  For any interval $I=(a,b]$, $N(I) \sim \mathrm{Poisson}(\lambda |I|)$
2.  For any $n$ disjoint intervals $I_1, I_2, \ldots, I_n$, the random variables
$N(I_1), N(I_2), \ldots, N(I_n)$ are independent.
:::

Note that property one implies that a homogeneous Poisson process has stationary
increments, alongside the independent increments from property two.

From this definition we get 
$$
\mathbb{P}(N(t) = n) = \frac{(\lambda t)^n e^{-\lambda t}}{n!}, \quad 
n=0,1,2,\ldots
$$ {#eq-pois_dist}

From @def-hpp, we can derive the properties of the process over an infinitesimal
interval $h>0$. Let $I=(t,t+h]$, so $|I|=h$. Using the Taylor expansion 
$e^{-\lambda h} = 1 - \lambda h + o(h)$, we can derive the following properties 
as $h \to 0$

1. $$ \begin{aligned} 
        \mathbb{P}(N(t+h)-N(t)=0) &= e^{-\lambda h} \\ 
        &= 1 - \lambda h + o(h) 
      \end{aligned}
   $$
2. $$ \begin{aligned}
        \mathbb{P}(N(t+h)-N(t)=1) &= \lambda h e^{-\lambda h} \\
        &= \lambda h(1 - \lambda h + o(h)) \\
        &= \lambda h + o(h) 
      \end{aligned}
   $$
3. $$ \begin{aligned} 
        \mathbb{P}(N(t+h)-N(t)\geq 2) &= 1 - \mathbb{P}(N(t+h)-N(t)=0) - \mathbb{P}(N(t+h)-N(t)=1) \\
        &= 1 - (1 - \lambda h + o(h)) - (\lambda h + o(h)) \\
        &= o(h)
      \end{aligned}
   $$

::: {#def-nhpp .definition}
## Nonhomogeneous Poisson Process

A counting process $(N(t): t \geq 0)$ is a nonhomogeneous Poisson process with
intensity function $\lambda(t) > 0$ if

1.  For any interval $I=(a,b]$,
$N(I) \sim \mathrm{Poisson}(\int_a^b \lambda(s) d s)$
2.  For any $n$ disjoint intervals $I_1, I_2, \ldots, I_n$, the random variables
$N(I_1), N(I_2), \ldots, N(I_n)$ are independent.
:::

Like we did for the homogeneous Poisson process, we can derive the infinitesimal
properties of the nonhomogeneous Poisson process. First we note that by the 
Fundamental Theorem of Calculus 
$\int_t^{t+h} \lambda(s) d s = \lambda(t)h + o(h)$, and the Taylor expansion 
$e^{-\lambda(t) h} = 1 - \lambda(t) h + o(h)$. Then, as $h \to 0$:

1. $$ \begin{aligned} 
        \mathbb{P}(N(t+h)-N(t)=0) &= e^{-\lambda(t) h + o(h)} \\
        &= 1 - \lambda(t) h + o(h) 
      \end{aligned}
   $$
2. $$ \begin{aligned}
        \mathbb{P}(N(t+h)-N(t)=1) &= (\lambda(t)h + o(h)) e^{-\lambda(t)h + o(h)} \\ 
        &= (\lambda(t)h + o(h))(1 - \lambda(t)h + o(h)) \\ 
        &= \lambda(t) h + o(h) 
      \end{aligned}
   $$
3. $$ \begin{aligned} 
        \mathbb{P}(N(t+h)-N(t)\geq 2) &= 1 - \mathbb{P}(N(t+h)-N(t)=0) - \mathbb{P}(N(t+h)-N(t)=1) \\
        &= 1 - (1 - \lambda(t)h + o(h)) - (\lambda(t)h + o(h)) \\
        &= o(h)
      \end{aligned}
   $$

### Conditional Intensity Functions

Given the history up until the last arrival $u$, $\mathcal{H}(u)$, define the 
conditional cumulative distribution (CDF) function of the next arrival time 
$T_{k+1}$ as 
$$
F(t \mid \mathcal{H}(u)) = \int_u^t \mathbb{P}(T_{k+1} \in [s, s + d s] 
\mid \mathcal{H}(u)) \, d s = \int_u^t f(s \mid \mathcal{H}(u)) \, d s
$$

The joint PDF for realisations $\{ t_1, t_2, \ldots, t_k \}$ is then, by the 
chain rule for probabilities $$
f(t_1, t_2, \ldots, t_k) = \prod_{i=1}^k f(t_i \mid \mathcal{H}(t_{i-1}))
$$ {#eq-joint_pdf}

It's very common to suppress notation of conditioning upon $\mathcal{H}(u)$ 
using a superscript asterisk notation, e.g. 
$f^*(t) \coloneqq f(t \mid \mathcal{H}(u))$.

It is often more convenient to work with the conditional intensity function 
(also known as the hazard function in some fields) $\lambda^*(t)$ defined as 
$$
\lambda^*(t) = \frac{f^*(t)}{1 - F^*(t)}
$$ {#eq-hazard}

However there is another representation for the conditional intensity as the 
expected rate of arrivals conditioned on $\mathcal{H}(t)$:

::: {#def-conditional_intensity .definition}
## Conditional Intensity Function

Consider a counting process $N$ with associated histories $\mathcal{H}$. If a 
non-negative function $\lambda^*(t)$ exists such that 
$$
\lambda^*(t) = \lim_{h \downarrow 0} \frac{\mathbb{E}[N(t + h) - N(t) 
\mid \mathcal{H}(t)]}{h}
$$ which only relies of information of $N$ in the past (that is, 
$\lambda^*(t)$ is $\mathcal{H}(t)$-measurable), then it is called the 
conditional intensity function of $N$.
:::

We note that $\lambda^*(t)$ is interpreted as its left-continuous modification
$\lambda^*(t^-)$ to ensure predictability of the counting process $N(\cdot)$.

Lastly, we define the compensator, which is frequently used in parameter
estimation and goodness of fit testing of point processes.

::: {#def-compensator .definition}
## Compensator

For a counting process $N$ the non-decreasing function 
$$
\Lambda(t) = \int_0^t \lambda^*(s) d s
$$ is called the compensator of the counting process.
:::

## Hawkes Process Background

The Hawkes process is a model for a ‘self-exciting’ counting process, where the 
history of past arrivals increases the chance of subsequent arrivals for a
period of time.

::: {#def-hawkes_process .definition}
## Hawkes Process

Consider $(N(t): t \geq 0)$ a counting process, with associated history 
$(\mathcal{H}(t): t \geq 0)$, that satisfies

$$
\mathbb{P}(N(t+h) - N(t) = m \mid \mathcal{H}(t)) = \begin{cases}
  1 - \lambda^*(t)h + o(h) &,m=0 \\
  \lambda^*(t)h + o(h) &,m=1 \\
  o(h) &,m\geq2 
\end{cases}
$$ {#eq-hp_definition} 

Suppose the process' conditional intensity function is of the form 
$$
\lambda^*(t) = \lambda + \int_0^t \mu(t - u) d N(u)
$${#eq-hp_lambda_integral} 
for some $\lambda > 0$ and $\mu: (0,\infty) \to [0, \infty]$ which are called 
the *background intensity* and *excitation function* respectively. Assume that 
$\mu(\cdot) \neq 0$ to avoid the trivial case (a homogeneous Poisson process). 
Such a process $N(\cdot)$ is a Hawkes process.
:::

Evaluating the stochastic integral in @eq-hp_lambda_integral, we obtain the more
commonly used version of the Hawkes process' conditional intensity function:
$$
\lambda^{*}(t) = \lambda + \sum_{T_i<t} \mu (t - T_i)
$$ {#eq-hp_lambda} 
where  $\{T_1,T_2,\ldots\}$ are all the arrival times in the history of 
the process $\mathcal{H}(t)$.

The most common choice of $\mu$ is the exponentially decaying kernel 
$$
\mu(t) = \alpha e^{-\beta t} , \quad \alpha,\beta > 0
$$ {#eq-hp_exp_kernel}

An alternative way to represent a Hawkes process is through a branching process
known as an immigrant-birth process (@Hawkes1974_ClusterRepresentation). In this
view, the immigration process is a homogeneous Poisson process with rate 
$\lambda$, and the arrival of an individual from the immigrant process at time 
$s$ initiates a non-homogeneous Poisson process of births with intensity 
$\mu(t - s)$ for $t > s$, which subsequently may result in further descendants. 
Each arrival generates a Poisson-distributed number of first-generation 
offspring with mean $\eta$, where $\eta$ is known as the branching ratio and is 
given by

$$
\eta = \int_0^{\infty}\mu(t)dt
$$ {#eq-hp_branchingratio}

For the process to be stationary with finite mean, the condition $\eta < 1$ is 
necessary and sufficient.

## Simulation

::: {#prp-lewis .proposition}
## Lewis' Thinning Algorithm

Suppose that events are occurring according to a Poisson process having rate
$\lambda$, and suppose that, independently of anything that came before, an event
that occurs at time $t$ is counted with probability $p(t)$. Then the process of
counted events, $N_c(t)$, constitutes a nonhomogeneous Poisson process with 
intensity function $\lambda(t) = \lambda p(t)$.
:::

::: {.proof}
The proposition is proved by confirming that the process $N_c(t)$ satisfies

a) $N(0) = 0$
b) The number of events that occur in disjoint time intervals are independent
c) $\mathbb{P}(N(t+h)-N(t)=1) = \lambda(t)h + o(h)$
d) $\mathbb{P}(N(t+h)-N(t)\geq 2) = o(h)$

As all events (not just the counted events) satisfy (a), (b), and (d), and we
count events independently, the corresponding result holds for the process
$N_c(t)$.

For (c), first we define for brevity $\Delta N(t,h) = N(t+h)-N(t)$. Then, by the
law of total probability
$$
\begin{aligned}
\mathbb{P}(\Delta N_c(t,h)) &= \mathbb{P}(\Delta N_c(t,h)=1 \mid \Delta N(t,h)=0) \times \mathbb{P}(\Delta N(t,h)=0) \\
&\quad + \mathbb{P}(\Delta N_c(t,h)=1 \mid \Delta N(t,h)=1) \times \mathbb{P}(\Delta N(t,h)=1) \\
&\quad + \mathbb{P}(\Delta N_c(t,h)=1 \mid \Delta N(t,h)\geq 2) \times \mathbb{P}(\Delta N(t,h)\geq 2) \\
&= 0 + (p(t) + o(1))(\lambda h + o(h)) + O(1)o(h) \\
&= \lambda p(t) + o(h)
\end{aligned}
$$
Therefore $N_c(t)$ is a Poisson process with intensity 
$\lambda p(t)$ on $[0, T]$

:::

Thus to generate the desired nonhomogeneous Poisson process, one can simulate 
from a homogeneous Poisson process with rate $\tilde{\lambda} \geq \lambda(t)$ 
for all $t \in [0, T]$, and then thin this process with 
$p(t) = \lambda(t) / \tilde{\lambda}$.


::: {#prp-ogata .proposition}
## Ogata's Thinning Construction for a Hawkes Process
Suppose candidate events occur according to a nonhomogeneous Poisson process 
with intensity $M(t)$ which is a predictable process satisfying 
$\lambda(t \mid \mathcal{H}(t)) \leq M(t)$ for all $t \in [0, T]$. Each 
candidate event occurring at time $t$ is accepted independently with probability

$$
p(t \mid \mathcal{H}(t)) = \frac{\lambda(t \mid \mathcal{H}(t))}{M(t)}.
$$

Then the process of accepted events $N_c(t)$ is a point process with conditional
intensity $\lambda(t \mid \mathcal{H}(t))$ on $[0, T]$.

:::

::: {.proof}

We verify that $N_c(t)$ satisfies the defining infinitesimal properties of a 
point process with conditional intensity $\lambda(t \mid \mathcal{H}(t))$:

a. $N_c(0) = 0$
b. $\mathbb{P}(\Delta N_c(t,h)=1 \mid \mathcal{H}(t)) = \lambda(t \mid \mathcal{H}(t))h + o(h)$
c. $\mathbb{P}(\Delta N_c(t,h)\ge2 \mid \mathcal{H}(t)) = o(h)$

Properties (a) and (c) follow directly from the corresponding properties of the 
candidate Poisson process.

The result for (b) is obtained by the law of total probability,

$$
\begin{aligned}
\mathbb{P}(\Delta N_c(t,h)=1 \mid \mathcal{H}(t))
&=
\mathbb{P}(\Delta N_c(t,h)=1 \mid \Delta N(t,h)=0,\mathcal{H}(t))
\mathbb{P}(\Delta N(t,h)=0) \\
&\quad+
\mathbb{P}(\Delta N_c(t,h)=1 \mid \Delta N(t,h)=1,\mathcal{H}(t))
\mathbb{P}(\Delta N(t,h)=1) \\
&\quad+
\mathbb{P}(\Delta N_c(t,h)=1 \mid \Delta N(t,h)\ge2,\mathcal{H}(t))
\mathbb{P}(\Delta N(t,h)\ge2) \\
&= 0 + \left(\frac{\lambda(t \mid \mathcal{H}(t))}{M(t)} + o(1)\right)(M(t) h + o(h)) + O(1)o(h) \\
&= \lambda(t \mid \mathcal{H}(t))h + o(h)
\end{aligned}
$$

so $N_c(t)$ has conditional intensity $\lambda(t \mid \mathcal{H}(t))$.

:::

::: {.remark}

In Ogata's algorithm the dominating rate $M(t)$ is implemented as a piecewise constant bound $M_n$ between successive accepted events. Candidate arrivals are generated from a homogeneous Poisson process with rate $M_n$, and the bound is updated whenever a new event occurs.

:::

## Maximum Likelihood Estimation

<!-- #### Example: Exponential MLE  {.unnumbered} -->

<!-- Below is an example of estimating a parameter from $\mathrm{Exp(\lambda)}$ data -->
<!-- using Maximum Likelihood Estimation (MLE). For this problem -->
<!-- $$ -->
<!-- \begin{aligned} -->
<!-- L(\lambda \mid \mathbf{t}) &= \prod_{i=1}^n \lambda e^{-\lambda t_i} \\ -->
<!-- \ell(\lambda \mid \mathbf{t}) &= n \log \lambda - \lambda \sum_{i=1}^n t_i -->
<!-- \end{aligned} -->
<!-- $$ -->

<!-- ```{r exp_mle} -->
<!-- # generate Exp(λ) variables, with each column a new realisation -->
<!-- generate_exp_matrix <- function(n_obs, n_trials, lambda_true, seed = NULL) { -->
<!--   # set seed if given -->
<!--   if (!is.null(seed)) {set.seed(seed)} -->

<!--   # simulate matrix of exponential samples -->
<!--   matrix( -->
<!--     rexp(n_obs*n_trials, rate = lambda_true), -->
<!--     nrow = n_obs, -->
<!--     ncol = n_trials -->
<!--   ) -->
<!-- } -->

<!-- # log-likelihood of Exp(λ) data -->
<!-- exp_loglik <- function(lambda, t) { -->
<!--   n <- length(t) -->
<!--   n*log(lambda) - lambda*sum(t) -->
<!-- } -->

<!-- # since optim minimises by default -->
<!-- exp_negloglik <- function(lambda, t) { -->
<!--   -exp_loglik(lambda, t) -->
<!-- } -->

<!-- # take t matrix and create a dataframe of mle parameter estimates and std errors -->
<!-- mle_results <- function(t_mat) { -->
<!--   n <- ncol(t_mat) -->

<!--   # to store results -->
<!--   results <- data.frame( -->
<!--     trial = 1:n,  -->
<!--     sample_size = rep(nrow(t_mat), n), -->
<!--     mean = rep(NA, n), -->
<!--     se = rep(NA, n), -->
<!--     lower = rep(NA, n), -->
<!--     upper = rep(NA, n) -->
<!--   ) -->

<!--   # mle fit for each trial -->
<!--   for (i in 1:n) { -->
<!--     fit <- optim( -->
<!--       par = 1, -->
<!--       fn = exp_negloglik, -->
<!--       t = t_mat[, i], -->
<!--       method = "L-BFGS-B", -->
<!--       lower = 1e-8, -->
<!--       hessian = TRUE -->
<!--     ) -->
<!--     results[i, "mean"] <- fit$par -->
<!--     results[i, "se"] <- sqrt(1 / fit$hessian)  -->
<!--   } -->
<!--   results$lower <- results$mean - 1.96*results$se -->
<!--   results$upper <- results$mean + 1.96*results$se -->

<!--   results -->
<!-- } -->

<!-- lambda_true <- 2 -->
<!-- t_mat <- generate_exp_matrix( -->
<!--   n_obs = 100,  -->
<!--   n_trials = 500,  -->
<!--   lambda_true = lambda_true,  -->
<!--   seed = 5 -->
<!-- ) -->
<!-- results <- mle_results(t_mat) -->
<!-- ``` -->

<!-- ```{r exp_mle_results_table, echo=FALSE} -->
<!-- summary_table <- data.frame( -->
<!--   Parameter = "λ", -->
<!--   MC_Trials = nrow(results), -->
<!--   Sample_Size = max(results$sample_size), -->
<!--   Mean = mean(results$mean), -->
<!--   Bias = mean(results$mean) - lambda_true, -->
<!--   # SD = sd(results$mean), -->
<!--   SD = sqrt(mean((results$mean - mean(results$mean))^2)), -->
<!--   RMSE = sqrt(mean((results$mean - lambda_true)^2)), -->
<!--   Coverage_95 = mean(results$lower <= lambda_true & -->
<!--                   lambda_true <= results$upper), -->
<!--   MC_Error = sd(results$mean) / sqrt(nrow(results)) -->
<!-- ) -->

<!-- knitr::kable( -->
<!--   summary_table, -->
<!--   digits = c(NA, 0, 0, 4, 4, 4, 4, 4, 4), -->
<!--   col.names = c( -->
<!--     "Parameter",  -->
<!--     "MC Trials", -->
<!--     "Sample Size", -->
<!--     "Mean",  -->
<!--     "Bias",  -->
<!--     "SD",  -->
<!--     "RMSE",  -->
<!--     "95% Coverage",  -->
<!--     "MC Error" -->
<!--   ) -->
<!-- ) -->
<!-- ``` -->


<!-- ```{r exp_mle_plot, echo=FALSE} -->
<!-- cover <- results$lower <= 2 & results$upper >= 2 -->
<!-- cols <- ifelse(cover, NORD_SLATE, NORD_RED) -->
<!-- widths <- ifelse(cover, 0.5, 1.5)  -->

<!-- plot_base_canvas( -->
<!--   xlim = c(min(results$trial), max(results$trial)),  -->
<!--   ylim = range(results$lower, results$upper), -->
<!--   title = "Coverage of 95% CI", -->
<!--   xlab = "Simulation", -->
<!--   ylab = expression(hat(lambda)) -->
<!-- ) -->
<!-- points(results$trial, results$mean, col = cols, pch = 16, cex = 0.6) -->
<!-- segments( -->
<!--   results$trial, results$lower, -->
<!--   results$trial, results$upper, -->
<!--   col = cols, lwd = widths -->
<!-- ) -->
<!-- abline(h = 2, lwd = 2) -->
<!-- ``` -->

::: {#thm-point_process_likelihood}
## Point Process Likelihood

Let $N$ be a simple point process with conditional intensity $\lambda^*(t)$ and
compensator $\Lambda(t)$. If we observe all the arrival times over the time 
period $[0, T ]$, denoted $\{t_1, \ldots, t_{N_T} \}$, then the likelihood 
function $L(\boldsymbol{\theta} \mid \mathbf{t})$ for $N$ is 
$$
L(\boldsymbol{\theta} \mid \mathbf{t})= \left[ \prod_{i=1}^{N_T} \lambda^*(t_i 
  \mid \boldsymbol{\theta}, \mathbf{t}) \right] 
  e^{-\Lambda(T \mid \boldsymbol{\theta}, \mathbf{t})}
$$ {#eq-pp_likelihood}
and the log-likelihood function is
$$
\ell(\boldsymbol{\theta} \mid \mathbf{t}) = \sum_{i=1}^{N_T} \log(\lambda^*(t_i \mid
  \boldsymbol{\theta}, \mathbf{t})) - \Lambda(T \mid \boldsymbol{\theta}, \mathbf{t})
$$ {#eq-pp_loglikelihood}
:::

For brevity we will not keep writing the observation vector $\mathbf{t}$ and the
parameter vector $\boldsymbol{\theta}$, so in that notation 
$L = L(\boldsymbol{\theta} \mid \mathbf{t})$, 
$\ell = \ell(\boldsymbol{\theta} \mid \mathbf{t})$, 
$\lambda^*(t_i) = \lambda^*(t_i \mid \boldsymbol{\theta}, \mathbf{t})$, and
$\Lambda(t) = \Lambda(t \mid \boldsymbol{\theta}, \mathbf{t})$.


#### Example: Poisson Process MLE  {.unnumbered}

For a simple example we'll estimate a Poisson process' parameter using MLE and
the form of the log-likelihood in @eq-pp_loglikelihood. In this case, 
$\lambda^*(t) = \lambda$ and the compensator $\Lambda(t) = \lambda t$. Therefore
$$
\ell = N_T \log \lambda - \lambda T
$$

The value of $\lambda$ that maximises this is easily found by setting 
$\frac{d \ell}{d \lambda}=0$ and solving for $\lambda$, giving 
$$
\hat{\lambda} = N_T/T
$$

And the code implementation estimating the parameters of a Poisson process 
observed over $[0, 10]$ with $\lambda = 2$:

```{r poissonprocess_mle}
lambda <- 2
T_max <- 10
set.seed(6)

# simulate a Poisson process over [0, T_max] using custom library
sim_object <- sim_pp(T_max, lambda)
plot(sim_object)
arrival_times <- sim_object$events

# log-likelihood of Poisson process arrival time data
poisprocess_loglik <- function(lambda, arrival_times, T_max) {
  N_T <- length(arrival_times)
  N_T*log(lambda) - lambda*T_max
}

# since optim minimises by default
poisprocess_negloglik <- function(lambda, arrival_times, T_max) {
  -poisprocess_loglik(lambda, arrival_times, T_max)
}

fit <- optim(
  par = 1, 
  fn = poisprocess_negloglik, 
  arrival_times = arrival_times,
  T_max = T_max,
  method = "L-BFGS-B",
  lower = 1e-8,
  hessian = TRUE
)

cat(sprintf(
  "Analytic Solution = %.3f ± %.3f \nNumeric MLE = %.3f ± %.3f", 
  length(arrival_times)/T_max, 1.96*sqrt(length(arrival_times))/T_max, 
  fit$par, 1.96/sqrt(fit$hessian)
))
```



### Hawkes Process MLE
To get the form of the Hawkes likelihood, first note that the integral in 
$\Lambda(T)$ over $[0, T]$ can be broken down into 
$$
[0, t_1] \cup (t_1,t_2] \cup \ldots \cup (t_{N_T - 1}, t_{N_T}] 
  \cup (t_{N_T}, T]
$$ 
and therefore 
$$
\Lambda(T) = \int_0^{t_i} \lambda^*(s) d s + \sum_{i=1}^{N_T-1} 
  \int_{t_i}^{t_i + 1} \lambda^*(s) d s + \int_{t_{N_T}}^{T} \lambda^*(s) d s
$$

Substituting the Hawkes conditional intensity in @eq-hp_lambda, and using 
$M(t) \coloneqq \int_0^t \mu(s) d s$, the integral term can be written as 
$$
\begin{aligned}
\int_{t_i}^{t_{i+1}} \lambda^*(s) d s &= \int_{t_i}^{t_{i+1}} 
  \left[ \lambda + \sum_{t_j < s} \mu(s - t_j) \right] d s 
  \\
&= \lambda (t_{i+1} - t_i) + \sum_{j=1}^i \int_{t_i}^{t_{i+1}} \mu(s - t_j) 
  d s 
  \\
&= \lambda (t_{i+1} - t_i) + \sum_{j=1}^i \left[ M(t_{i+1} - t_j) - 
  M(t_i - t_j) \right]
\end{aligned}
$$

This summation holds for $i=1,\ldots,N_T-1$, for $i=0$ where the summation term
disappears due to no events in the history, and for $i = N_T$ if we abuse 
notation a little and say $T=t_{N_T+1}$. Therefore we can write
$$
\begin{aligned}
\Lambda(T) &= \sum_{i=1}^{N_T} \left\{ \lambda(t_{i+1} - t_i) + 
  \sum_{j=1}^i [M(t_{i+1} - t_j) - M(t_i - t_j)] \right\} 
  \\ 
&= \lambda T + \sum_{i=1}^{N_T} \sum_{j=1}^i [M(t_{i+1} - t_j) - M(t_i - t_j)]
  \\
&= \lambda T + \sum_{i=1}^{N_T}  [M(T - t_i) - M(0)] 
  = \lambda T + \sum_{i=1}^{N_T}  M(T - t_i)
\end{aligned}
$$ {#eq-hawkes_compensator}

The log-likelihood (@eq-pp_loglikelihood), combined with the this form of the 
Hawkes compensator $\Lambda(T)$ and the Hawkes conditional intensity 
(@eq-hp_lambda) gives us

$$
\ell = \sum_{i=1}^{N_T} \log \left[ \lambda + \sum_{j=1}^{i-1} \mu(t_i - t_j)  
  \right] - \lambda T - \sum_{i=1}^{N_T}  M(T - t_i)
$$
The double sum will mean that for a numeric optimisation routine, we will have 
to evaluate a $\mathcal{O}([N(T)]^2)$ function for each optimisation step which 
will scale poorly.

#### Example: Exponential Hawkes Process MLE  {.unnumbered}

For a Hawkes process with an exponential kernel (@eq-hp_exp_kernel), the 
log-likelihood function is
$$
\ell = \sum_{i=1}^{N_T} \ln \left[ \lambda + \alpha \sum_{j=1}^{i-1} 
  e^{-\beta (t_i - t_j)} \right] - \Lambda (T)
$${#eq-exp_hp_loglik}
where the compensator of the process is, using @eq-hawkes_compensator, 
$$
\Lambda(t) = \lambda T + \sum_{i=1}^{N_T}M(T-t_i) 
  = \lambda T + \frac{\alpha}{\beta} \sum_{i=1}^{N_T} 
  \left[ 1 - e^{-\beta (T - t_i)} \right]
$$

The inner sum in @eq-exp_hp_loglik can simplified for faster computation. In 
particular, for $i \in \left\{2,...,N_T \right\}$, let 
$A(i) = \sum_{j=1}^{i-1} e^{-\beta(t_i - t_j)}$ (the inner sum) so that we can 
find the recursive relation
$$
\begin{aligned}
A(i) &= \sum_{j=1}^{i-1} e^{-\beta(t_i - t_j)} = \sum_{j=1}^{i-1} 
  e^{-\beta(t_i - t_j + t_{i-1} - t_{i-1})} \\
&= e^{-\beta(t_i - t_{i-1})} \sum_{j=1}^{i-1} e^{-\beta(t_{i-1} - t_j)} 
  = e^{-\beta(t_i - t_{i-1})}\left(1 +  \sum_{i=1}^{j-2} 
  e^{-\beta(t_{i-1} - t_j)}  \right) \\
&= e^{-\beta(t_i - t_{i-1})}(1 + A(i-1))
\end{aligned}
$$

With the base case that $A(1) = 0$, the exponential Hawkes likelihood 
(@eq-exp_hp_loglik) can be rewritten
$$
\ell = \sum_{i=1}^{N_T} \ln \left[ \lambda + \alpha A(i) \right] - \Lambda (T)
$$ {#eq-exp_hp_loglik_fast}

And the code implementation estimating the parameters of an exponential Hawkes 
process over $[0, 200]$ with 
$\boldsymbol{\theta} = (\lambda, \alpha, \beta) = (0.5, 2, 3)$:
```{r hawkesprocess_mle}
exp_hp_intensity <- function(theta, t, H_t) {
  # unpack parameters
  lambda <- theta[1]
  alpha <- theta[2]
  beta <- theta[3]
  
  lambda + alpha*sum(exp(-beta*(t - H_t)))
}

exp_hp_compensator <- function(theta, t, H_t) {
  # unpack parameters
  lambda <- theta[1]
  alpha <- theta[2]
  beta <- theta[3]
  
  if (t <= 0) return(0)
  lambda*t + (alpha/beta)*sum(1 - exp(-beta*(t - H_t)))
}

exp_hp_loglik <- function(theta, T_max, H_t) {
  # there is a speedup for the exponential hawkes log-likelihood that is O(N(T_max))
  # this is the more expensive O([N(T_max)]^2) computation
  log_intensities <- sapply(H_t, function(t_i) {
    past_events <- H_t[H_t < t_i] # Find history by comparing values
    log(exp_hp_intensity(theta = theta, t = t_i, H_t = past_events))
  })
  
  # total log-likelihood = sum(log(intensities)) - compensator at T
  sum(log_intensities) - exp_hp_compensator(theta = theta, t = T_max, H_t = H_t)
}

exp_hp_loglik_fast <- function(theta, T_max, H_t) {
  lambda <- theta[1]
  alpha <- theta[2]
  beta <- theta[3]
  
  n <- length(H_t)
  # Recursive computation of the summation term A(i)
  A <- numeric(n)
  if (n > 1) {
    for (i in 2:n) {
      A[i] <- exp(-beta * (H_t[i] - H_t[i-1])) * (1 + A[i-1])
    }
  }
  
  log_intensities <- log(lambda + alpha*A)
  sum(log_intensities) - exp_hp_compensator(theta, T_max, H_t)
}

exp_hp_negloglik<- function(theta, T_max, H_t, fast=TRUE) {
  if (fast) return(-exp_hp_loglik_fast(theta = theta, T_max = T_max, H_t = H_t))
  else return(-exp_hp_loglik(theta = theta, T_max = T_max, H_t = H_t))
}

T_max <- 200
lambda <- 0.5
alpha <- 2
beta <- 3;

set.seed(123)
sim_object <- sim_hp(
  T_max = T_max, 
  lambda = lambda, 
  mu = function(t) {alpha*exp(-beta*t)}
)
H_t <- sim_object$events

fit <- optim(
  par = c(1, 1, 1), 
  fn = exp_hp_negloglik, 
  H_t = H_t,
  T_max = T_max,
  method = "L-BFGS-B",
  lower = c(1e-6, 1e-6, 1e-6),
  hessian = TRUE
)

mle <- fit$par
v_cov <- solve(fit$hessian)
se <- sqrt(diag(v_cov))

cat("Maximum Likelihood Estimates with 95% CIs:\n", sprintf(
  "lambda_mle = %.3f ± %.3f \n alpha_mle = %.3f ± %.3f \n beta_mle = %.3f ± %.3f", 
  mle[1], 1.96*se[1], 
  mle[2], 1.96*se[2], 
  mle[3], 1.96*se[3]
))

```

```{r hawkesprocess_mle_plot}
t_grid <- seq(0, T_max, length.out = 2000) 
intensities <- sapply(t_grid, function(t_i) {
  past <- H_t[H_t < t_i]
  lambda + sum(alpha*exp(-beta*(t_i - past)))
})

plot_base_canvas(c(0, T_max), c(0, max(intensities)*1.1), "Hawkes MLE Fit")
abline(h = fit$par[1], col = NORD_NIGHT, lty = 3) # Baseline
add_intensity(t_grid, intensities)
add_events(H_t)
```


## Marked Hawkes Processes

Let $M_i \in \mathcal{M}$ be the mark associated with arrival time $T_i$. Let us
say that $N([s,t), \mathcal{A})$ counts the number of arrivals with marks 
$\mathcal{A} \subset \mathcal{M}$ from time $s$ to time $t$. The the definition
for the conditional intensity process becomes
$$
\lambda^*(t,m) = \lim_{\Delta t\,\Delta m\, \downarrow 0} = 
  \frac{\mathbb{E}[N([t, t + \Delta t), [m + \Delta m) ) \mid \mathcal{H}_t]}
  {|\Delta m| |\Delta t|}
$$
if this limit exists (now $\mathcal{H}_t$ id the filtration generated by 
$\{(T_i,M_i):T_i<t \}$). Typicall this is written as
$$
\lambda^*(t,m) = \lambda_g^*(t)f^*(m \mid t)
$$ {#eq-hp_marked_lambda} 
where $\lambda_g^*(t)$ is the intensity of the underlying 'ground' point process
(conditioned on $\mathcal{H}_t$ so may depend on history of marks), and 
$f^*(m \mid t)$ is the density of the mark $m$ at time $t$ (also conditioned on 
$\mathcal{H}_t$). The likelihood function is then
$$
\begin{aligned}
\mathcal{L}(\boldsymbol{\theta} \mid \mathcal{H}_T,T) &= 
  \left[\prod_{i=1}^n \lambda^*(t_i, m_i)\right] \times \exp \left(-\int_0^T 
  \int_{\mathcal{M}} \lambda^*(s,m) d m \, d s \right) \\
&= \left[ \prod_{i=1}^n \lambda_g^*(t_i) f^*(m_i \mid t_i) \right] \times 
  \exp \left( -\int_0^T \int_{\mathcal{M}} \lambda_g^*(s) f^*(m \mid s) 
  d m \, d s \right) \\
&= \left[ \prod_{i=1}^n \lambda_g^*(t_i) f^*(m_i \mid t_i) \right] \times 
  \exp \left( -\int_0^T \lambda_g^*(s) d m \, d s \right)
\end{aligned}
$$ {#eq-hp_marked_lambda_likelihood} 
and the log-likelihood is
$$
\ell(\boldsymbol{\theta} \mid \mathcal{H}_T,T) = \sum_{i=1}^n \log \lambda_g^*(t_i)
  + \sum_{i=1}^n \log f^*(m_i \mid t_i) - \int_0^T \lambda_g^*(s) d s
$$ {#eq-hp_marked_lambda_loglikelihood} 

The marks are called unpredictable if the marks are independent of the history 
($f^*(m \mid t) = f(m \mid t)$) though they may impact the underlying Hawkes 
intensity process. 

For unpredictable marks:
$$
\begin{aligned}
\ell(\boldsymbol{\theta} \mid \mathcal{H}_T,T) &= \sum_{i=1}^n \log \lambda_g^*(t_i)
  - \int_0^T \lambda_g^*(s) d s + \sum_{i=1}^n \log f(m_i \mid t_i)  \\
&= \ell_{\lambda}(\boldsymbol{\theta}_{\lambda} \mid \mathcal{H}_T,T) 
  + \ell_f(\boldsymbol{\theta}_f \mid \mathcal{H}_T,T)
\end{aligned}
$$ {#eq-hp_marked_lambda_loglikelihood2} 
where $\ell_{\lambda}(\cdot)$ is the log-likelihood function for the ground intensity 
process, and $\ell_f(\cdot)$ is the log-likelihood function for the mark density.
$$
\begin{aligned}
\ell_{\lambda}(\boldsymbol{\theta}_{\lambda} \mid  \mathcal{H}_T,T) &= 
  \sum_{i=1}^n \log \lambda_g^*(t_i) - \int_0^T \lambda_g^*(s) d s \\
\ell_f(\boldsymbol{\theta}_f \mid  \mathcal{H}_T,T) &= 
  \sum_{i=1}^n \log f(m_i \mid t_i)
\end{aligned}
$$
$\boldsymbol{\theta}_{\lambda}$ is the parameter vector for the ground intensity process and
$\boldsymbol{\theta}_f$ is the parameter vector for the mark density. The benefit of the 
unpredictable mark assumption is that it allows this separable form of the 
log-likelihood function which means each component can be maximised separately.

### Example: NZ Quake Catalogue

The dataset was downloaded from [GeoNet Quake Search](https://quakesearch.geonet.org.nz/)
with the following parameters

- Earthquakes between 2000-01-01 00:00 UTC and 2025-01-01 00:00 UTC
- Magnitude threshold $m_0 = 4.0$
- Shallow events down to 40km
- Central New Zealand (the rectangular area 171°--179°E and 38°--43°S)

```{r load_plot_nz_quakes}
nz_quakes <- read.csv("data/nz_quakes_2000-2025_m4plus_shallow_central.csv")
start_time <- as.POSIXct("2000-01-01 00:00:00", tz = "UTC")
end_time <- as.POSIXct("2025-01-01 00:00:00", tz = "UTC")
T_max <- as.numeric(difftime(end_time, start_time, units = "days"))
mag_thresh <- 4

nz_quakes$origintime <- as.POSIXct(
  nz_quakes$origintime, 
  format = "%Y-%m-%dT%H:%M:%OSZ", 
  tz = "UTC"
)

nz_quakes$days_from_origin <- as.numeric(difftime(
  nz_quakes$origintime, 
  start_time, 
  units = "days"
))

nz_quakes <- sort_by(nz_quakes, ~days_from_origin)

plot_base_canvas(
  xlim  = c(0, T_max), 
  ylim  = c(0, nrow(nz_quakes) + 1), 
  title = "Central NZ Quake Catalogue: N(t)",
  ylab  = expression(N(t))
)
add_counting_process(nz_quakes$days_from_origin, T_max)
add_events(nz_quakes$days_from_origin)

```
```{r plot_nz_quake_map}
nz_map <- ne_countries(
  scale = "medium", 
  country = "New Zealand", 
  returnclass = "sf")

ggplot() +
  geom_sf(data = nz_map, color = "white") + 
  geom_point(
    data = nz_quakes, 
    aes(x = longitude, y = latitude, size = magnitude, color = magnitude), 
    alpha = 0.5) +
  scale_color_viridis_c() + 
  coord_sf(xlim = c(171, 179), ylim = c(-38, -43)) + 
  theme_minimal() +
  labs(
    title = "Shallow NZ Earthquakes (>M4, <40km deep)", 
    subtitle = "2000 - 2025", 
    size = "Magnitude", 
    color = "Magnitude"
  )
```

The model I'll attempt to fit is the class of Temporal ETAS models. These are 
marked Hawkes processes with ground intensity
$$
\lambda_g^*(t) = \lambda + \sum_{T_i<t} Ae^{\alpha(M_i-m_0)} \frac{p-1}{c} 
  \left(1+\frac{t - T_i}{c} \right)^{-p}
$$
for $\lambda,A,\alpha,c,m_0>0,p>1$, and mark density (iid)
$$
f(m) = \beta e^{-\beta(M-m_0)}
$$
for $\beta, m_0>0$. 

Therefore the log-likelihood is 
$$
\ell(\boldsymbol{\theta} \mid \mathcal{H}_T,T) = 
  \ell_{\lambda}(\boldsymbol{\theta}_{\lambda} \mid \mathcal{H}_T,T) +
  \ell_f(\boldsymbol{\theta}_f \mid \mathcal{H}_T,T),
$$
where
$$
\begin{aligned}
\ell_{\lambda}(\boldsymbol{\theta}_{\lambda} \mid \mathcal{H}_T,T) &= 
  \sum_{i=1}^n \log \lambda_g^*(t_i) - \int_0^T \lambda_g^*(s) ds, \\
\ell_f(\boldsymbol{\theta}_f \mid \mathcal{H}_T,T) &= 
  \sum_{i=1}^n \log f(m)
\end{aligned}
$$
which can be maximised independently.

For $\ell_f(\cdot)$, $m \sim \mathrm{Exp}(\beta)$ which has the MLE 
$\hat{\beta} = 1/\bar{m}$ with $\mathrm{SE}(\hat{\beta})=\hat{\beta}/\sqrt{n}$.

For the $\ell_{\lambda}(\cdot)$ term, we will want a closed form expression for
$\Lambda_g(t)=\int_0^T \lambda_g^*(s) ds$. For this we can use @eq-hawkes_compensator, 
where $M(t)$ is given by:

$$
\begin{aligned}
M(t) 
  &= \int_0^t A e^{\alpha(M_i-m_0)} \frac{p-1}{c} \left(1 + \frac{s}{c} \right)^{-p} {\rm d}s \\
  &= A e^{\alpha(M_i-m_0)} (p-1) \int_1^{1+\frac{t}{c}} u^{-p} {\rm d}u \\
  &= A e^{\alpha(M_i-m_0)} (p-1) \left[\frac{u^{1-p}}{1-p}\right]_1^{1+\frac{t}{c}} \\
  &= A e^{\alpha(M_i-m_0)} \left( 1 - \left(1 + \frac{t}{c}\right)^{1-p} \right)
\end{aligned}
$$

and therefore:

$$
\begin{aligned}
\Lambda_g(t) 
  &= \lambda t + \sum_{i=1}^{N_t}  M(t - t_i) \\
  &= \lambda t + A\sum_{T_i<t}e^{\alpha(M_i-m_0)}\left(1-\left(1+\frac{t-T_i}{c} \right)^{1-p} \right)
\end{aligned}
$$
This allows us to optimise $\ell_{\lambda}(\cdot)$ numercially using `optim`:

```{r fit_mhp_nz_quakes}
H_t <- data.frame(
  time = nz_quakes$days_from_origin, 
  mark = nz_quakes$magnitude - mag_thresh # simpler to work with shifted magnitudes
)

# simple exponential mark density (of shifted magnitude)
# therefore mle = 1/mean(marks)
beta <- 1/mean(H_t$mark)
beta_se <- beta/sqrt(length(H_t$mark))

# ground process terms (full history dependence)
etas_power_intensity <- function(theta, t, H_t) {
  lambda <- theta[1]
  A <- theta[2]
  alpha <- theta[3]
  c <- theta[4]
  p <- theta[5]
  
  lambda + (A*(p - 1)/c)*sum(exp(alpha*H_t$mark) * (1 + (t - H_t$time)/c)^(-p))
}

# etas_exp_intensity <- function(theta, t, H_t) {
#   lambda <- theta[1]
#   A <- theta[2]
#   alpha <- theta[3]
#   gamma <- theta[4]
#   
#   lambda + A*sum(exp(alpha*H_t$mark) * exp(-gamma*(t - H_t$time)))
# }

etas_power_compensator <- function(theta, t, H_t) {
  lambda <- theta[1]
  A <- theta[2]
  alpha <- theta[3]
  c <- theta[4]
  p <- theta[5]
  
  lambda*t + A*sum(exp(alpha*H_t$mark)*(1 - (1 + (t - H_t$time)/c)^(1-p)))
}

# etas_exp_compensator <- function(theta, t, H_t) {
#   lambda <- theta[1]
#   A <- theta[2]
#   alpha <- theta[3]
#   gamma <- theta[4]
#   
#   lambda*t + (A/gamma)*sum(exp(alpha*H_t$mark)*(1 - exp(-gamma*(t - H_t$time))))
# }

etas_power_loglik <- function(theta, T_max, H_t) {
  # this is an expensive O([N(T_max)]^2) computation
  log_intensities <- sapply(1:nrow(H_t), function(i) {
    t_i <- H_t$time[i]
    past_events <- H_t[H_t$time < t_i, ]
    log(etas_power_intensity(theta = theta, t = t_i, H_t = past_events))
  })
  
  # total log-likelihood = sum(log(intensities)) - compensator at T
  sum(log_intensities) - etas_power_compensator(theta = theta, t = T_max, H_t = H_t)
}

etas_power_negloglik <- function(theta, T_max, H_t) {
  -etas_power_loglik(theta, T_max, H_t)
}

# etas_exp_loglik <- function(theta, T_max, H_t) {
#   # this is an expensive O([N(T_max)]^2) computation
#   log_intensities <- sapply(1:nrow(H_t), function(i) {
#     t_i <- H_t$time[i]
#     past_events <- H_t[H_t$time < t_i, ]
#     log(etas_exp_intensity(theta = theta, t = t_i, H_t = past_events))
#   })
# 
#   # total log-likelihood = sum(log(intensities)) - compensator at T
#   sum(log_intensities) - etas_exp_compensator(theta = theta, t = T_max, H_t = H_t)
# }

# etas_exp_negloglik <- function(theta, T_max, H_t) {
#   -etas_exp_loglik(theta, T_max, H_t)
# }

fit_power <- optim(
  par = c(0.2, 0.2, 1.2, 0.2, 1.2),
  fn = etas_power_negloglik,
  H_t = H_t,
  T_max = T_max,
  method = "L-BFGS-B",
  lower = c(1e-6, 1e-6, 1e-6, 1e-6, 1+1e-6),
  control=list(trace=2),
  hessian = TRUE
)

mle_power <- fit_power$par
v_cov_power <- solve(fit_power$hessian)
se_power <- sqrt(diag(v_cov_power))

cat("\nMaximum Likelihood Estimates with 95% CIs:\n", sprintf(
  "beta_mle = %.3f ± %.3f \n lambda_mle = %.3f ± %.3f \n A_mle = %.3f ± %.3f \n alpha_mle = %.3f ± %.3f \n c_mle = %.3f ± %.3f \n p_mle = %.3f ± %.3f \n\n",
  beta, 1.96*beta_se,
  mle_power[1], 1.96*se_power[1],
  mle_power[2], 1.96*se_power[2],
  mle_power[3], 1.96*se_power[3],
  mle_power[4], 1.96*se_power[4],
  mle_power[5], 1.96*se_power[5]
))

# fit_exp <- optim(
#   par = c(1, 1, 1, 1), 
#   fn = etas_exp_negloglik, 
#   H_t = H_t,
#   T_max = T_max,
#   method = "L-BFGS-B",
#   lower = c(1e-6, 1e-6, 1e-6, 1e-6),
#   control=list(trace=2),
#   hessian = TRUE
# )

# mle_exp <- fit_exp$par
# v_cov_exp <- solve(fit_exp$hessian)
# se_exp <- sqrt(diag(v_cov_exp))

# cat("\nMaximum Likelihood Estimates with 95% CIs:\n", sprintf(
#   "beta_mle = %.3f ± %.3f \n lambda_mle = %.3f ± %.3f \n A_mle = %.3f ± %.3f \n alpha_mle = %.3f ± %.3f \n gamma_mle = %.3f ± %.3f \n\n", 
#   beta, 1.96*beta_se,
#   mle_exp[1], 1.96*se_exp[1], 
#   mle_exp[2], 1.96*se_exp[2], 
#   mle_exp[3], 1.96*se_exp[3], 
#   mle_exp[4], 1.96*se_exp[4]
# ))
```

## Goodness of Fit

A useful technique exists to assess the goodness-of-fit of some point data to 
a generic model for the point data. This technique uses the compensator, 
$\Lambda(t)$, of the model, and the random time change theorem of 
@DaleyVereJones2003_PointProcessVol1 Proposition 7.4.IV:

::: {#thm-random_time_change .theorem}
## Random Time Change and Residual Analysis
Consider an unbounded, increasing sequence of time points $\{t_1,t_2,\ldots\}$ 
in the half line $(0,\infty)$, and a monotone, continuous compensator 
$\Lambda(\cdot)$ such that $\lim_{t\to\infty} \Lambda(t)=\infty$ almost surely.
The transformed sequence $\{t_1^*,t_2^*,\ldots\}=\{\Lambda(t_1),\Lambda(t_2),\ldots\}$,
where the counting process is denoted $N^*(t)$ is a realisation of a unit rate 
Poisson process if and only if the original sequence $\{t_1,t_2,\ldots\}$ is a 
realisation from the point process defined by $\Lambda(\cdot)$.

:::

Hence, equipped with a closed form of the compensator, the quality of the
inference can be ascertained using standard fitness tests for Poisson processes.

This method is applied to the power decay Hawkes models fit of the central NZ 
earthquake catalogue. We compare the quantiles of the transformed process to the
theoretical values of a $\mathrm{Exp(1)}$ random variable, and perform a
Kolmogorov-Smirnov test.

Below we can see that the power more closely follows the theoretical quantiles, 
and the KS test agrees with us not rejecting the null hypothesis at a 5% level 
that the transformed process follows a $\mathrm{Exp(1)}$ distribution.
Therefore we cannot reject the hypothesis that the original point process 
follows a Hawkes process defined by the compensator used to transform the data.

```{r}
# For each event i, calculate the compensator value at time t_i
theta <- fit_power$par
tau_power <- sapply(1:nrow(H_t), function(i) {
  ti <- H_t[i, "time"]
  hi <- H_t[1:(i - 1), ] # Only use history up to this point
  etas_power_compensator(theta = theta, H_t = hi, t = ti)
})

# convert to inter-arrival times 
gaps_power <- diff(c(0, tau_power))
n <- length(gaps_power)

max_val_power <- max(qexp(stats::ppoints(n)), gaps_power) * 1.05

plot_base_canvas(
  xlim  = c(0, max_val_power), 
  ylim  = c(0, max_val_power), 
  title = "Residual Q-Q Plot - Power Decay",
  xlab  = "Theoretical Exp(1) Quantiles",
  ylab  = "Sample Residual Quantiles"
)
graphics::abline(0, 1, col = "#BF616A", lty = 2, lwd = 2)
theoretical_q <- qexp(stats::ppoints(n))
graphics::points(theoretical_q, sort(gaps_power), col = "#5E81AC", pch = 20, cex = 0.7)

# subtitle for the Kolmogorov-Smirnov test
ks_res <- ks.test(gaps_power, "pexp", rate = 1)
ks_text <- sprintf("K-S Test p-value: %.4f", ks_res$p.value)
graphics::mtext(ks_text, side = 3, line = 0.2, cex = 0.8, col = "#4C566A")

```


```{r, include=FALSE}
# sim_mhp <- function(T_max, intensity_func, rmark_func, ...) {
#   
#   H_t = data.frame(time = 0, mark = 0)
#   t <- 0
#   count <- 0
#   
#   while (t < T_max) {
#     
#     # find new upper bound, using only the filled portion of H_t
#     M <- intensity_func(t = t, H_t = H_t[seq_len(count), ], ...)
#     # M <- lambda + sum(mu(t - arrival_times[seq_len(count)]))
#     if (is.na(M)) {
#       stop("Intensity calculation returned NA. Check your function.")
#     }
#     
#     # generate next candidate point
#     t <- t + rexp(1, M)
#     if (t > T_max) break
#     
#     current_intensity <- intensity_func(t = t, H_t = H_t[seq_len(count), ], ...)
#     # keep it with some probability
#     if (runif(1) <= current_intensity/M) {
#       count <- count + 1
#       mark <- rmark_func()
#       H_t[count, ] <- c(t, mark)
#     }
#   }
#   
#   return(H_t)
# }
# 
# set.seed(3141)
# rmark_func <- function() {rexp(1, beta)}
# H_t_sim <- sim_mhp(10000, etas_power_intensity, rmark_func, theta=fit_power$par)
# plot_base_canvas(
#   xlim  = c(0, 10000), 
#   ylim  = c(0, nrow(H_t_sim)), 
#   title = "Simulated Marked Hawkes Process: N(t)",
#   ylab  = expression(N(t))
# )
# add_counting_process(H_t_sim$time, 10000)
# add_events(H_t_sim$time)
# 
# fit_power_sim <- optim(
#   par = c(0.2, 0.2, 1.2, 0.2, 1.2), 
#   fn = etas_power_negloglik, 
#   H_t = H_t_sim,
#   T_max = 10000,
#   method = "L-BFGS-B",
#   lower = c(1e-6, 1e-6, 1e-6, 1e-6, 1+1e-6),
#   control=list(trace=2),
#   hessian = TRUE
# )
# 
# beta_sim <- 1/mean(H_t_sim$mark)
# beta_se_sim <- beta_sim/sqrt(nrow(H_t_sim))
# mle_power_sim <- fit_power_sim$par
# v_cov_power_sim <- solve(fit_power_sim$hessian)
# se_power_sim <- sqrt(diag(v_cov_power_sim))
# 
# cat("\nMaximum Likelihood Estimates with 95% CIs:\n", sprintf(
#   "beta_mle = %.3f ± %.3f \n lambda_mle = %.3f ± %.3f \n A_mle = %.3f ± %.3f \n alpha_mle = %.3f ± %.3f \n c_mle = %.3f ± %.3f \n p_mle = %.3f ± %.3f \n\n", 
#   beta_sim, 1.96*beta_se_sim,
#   mle_power_sim[1], 1.96*se_power_sim[1], 
#   mle_power_sim[2], 1.96*se_power_sim[2], 
#   mle_power_sim[3], 1.96*se_power_sim[3], 
#   mle_power_sim[4], 1.96*se_power_sim[4], 
#   mle_power_sim[5], 1.96*se_power_sim[5]
# ))
# 
# cat("Actual params = ", c(beta, fit_power$par))
```
## References {.unnumbered}
