---
title: "Marked Hawkes Process Gibbs Sampler"
author: "Devon Grossett"
format: 
  html:
    css: styles.css
    theme: cosmo
    code-fold: false
---

```{r, warning=FALSE}
library(HawkesProject)
```


The conditional intensity of a Marked Hawkes Process (MHP) can be expressed as
$$
\lambda^*(t,m) = \lambda_g^*(t) f^*(m \mid t)
$$
where $f^*(m \mid t)$ is the mark density function, and $\lambda_g^*(t)$ the 
ground process intensity function. The asterisk ($*$) indicates dependence of 
the functions on the process history up to time $t$, 
$\mathcal{H}(t) = \{(t_i, m_i): t_i < t \}$

The ground intensity process can be represented as
$$
\lambda_g^*(t) = \lambda + \sum_{t_i < t}h(t-t_i,m_i)
$$

I'll look at processes where the mark history enters through an exponential 
(typical in Hawkes models for earthquakes), and decay exponentially in time, 
e.g.
$$
\begin{gather}
h(t-t_i, m_i) = \eta(m_i)g(t-t_i) \\
\eta(m) = Ae^{\alpha(m)} \\
g(t) = \beta e^{-\beta t}
\end{gather}
$$
and an iid exponential mark density, e.g.
$$
f^*(m \mid t) = f(m) = \gamma e^{-\gamma t}
$$

A useful thing to know for estimation is the integrated (ground) intensity 
function $\Lambda_g(t)=\int_0^t \lambda_g^*(t) dt$. For this model, this is:
$$
\begin{aligned}
\Lambda_g(t)
&= \int_0^t \left( \lambda + \sum_i A e^{\alpha m_i} 
  \beta e^{-\beta(s-t_i)} \mathbf{1}_{\{t_i < s\}} \right) ds \\
&= \lambda t + \sum_{i} \int_0^t A e^{\alpha m_i} 
  \beta e^{-\beta(s-t_i)} \mathbf{1}_{\{t_i < s\}} \, ds \\
&= \lambda t + \sum_{t_i<t} \int_{t_i}^t A e^{\alpha m_i} 
  \beta e^{-\beta(s-t_i)} ds \\
&= \lambda t + \sum_{t_i<t} \int_0^{t - t_i} 
  A e^{\alpha m_i} \beta e^{-\beta u} \, du \\
&= \lambda t + \sum_{t_i<t} 
  A e^{\alpha m_i} \left(1 - e^{-\beta(t - t_i)}\right)
\end{aligned}
$$

Some code to simulate from a (MHP) $\{(t_1,m_1),\ldots,(t_n,m_n) \}$ observed 
over $[0, T_{\max}]$. 

```{r}
sim_mhp <- function(T_max, intensity_func, rmark_func, ...) {
  
  # initialise data frame. 1st row will be overwritten so values don't matter
  H_t = data.frame(time = 0, mark = 0)
  t <- 0
  count <- 0
  
  while (t < T_max) {
    # find new upper bound, using only the filled portion of H_t
    # only valid for intensities that are non-increasing between events
    M <- intensity_func(t = t, H_t = H_t[seq_len(count), ], ...)
    if (is.na(M)) {
      stop("Intensity calculation returned NA. Check function.")
    }
    
    # generate next candidate point
    t <- t + rexp(1, M)
    if (t > T_max) break
    
    process_intensity <- intensity_func(t = t, H_t = H_t[seq_len(count), ], ...)
    # keep it with probability process_intensity/M
    if (runif(1) <= process_intensity/M) {
      count <- count + 1
      mark <- rmark_func()
      H_t[count, ] <- c(t, mark)
    }
  }
  return(H_t)
}

# exponentially decaying kernel
exp_intensity <- function(theta, t, H_t) {
  lambda <- theta[1]
  A <- theta[2]
  alpha <- theta[3]
  beta <- theta[4]
  
  lambda + A*beta*sum(exp(alpha*H_t$mark - beta*(t - H_t$time)))
}

exp_compensator <- function(theta, t, H_t) {
  lambda <- theta[1]
  A <- theta[2]
  alpha <- theta[3]
  beta <- theta[4]
  
  lambda*t + A*sum(exp(alpha*H_t$mark)*(1 - exp(-beta*(t - H_t$time))))
}

```

Simulating a MHP with 
$(\lambda,A,\alpha,\beta,\gamma) = (0.1,\, 0.2,\, 2,\, 2.5,\, 2,2)$:

```{r}
T_max <- 2000
theta <- c(0.1, 0.08, 2, 2.5, 2.2)
branching_ratio <- (theta[2]/theta[4]) * (theta[5]/(theta[5] - theta[3]))

# e.g. earthquake magnitude
rmark_func <- function(gamma = theta[5]) {rexp(1, gamma)}

set.seed(1)
simulated_process <- sim_mhp(T_max, exp_intensity, rmark_func, theta = theta)

# My HawkesProject plotting utilities 
plot_base_canvas(
  xlim  = c(0, T_max), 
  ylim  = c(0, nrow(simulated_process) + 1), 
  title = expression(paste("Simulated MHP: ", N[g](t))),
  ylab  = expression(N[g](t))
)
add_counting_process(simulated_process$time, T_max)
add_events(simulated_process$time)
```

The log-likelihood for this process observed over $[0,T]$ is

$$
\begin{aligned}
\ell(\boldsymbol{\theta} \mid \mathcal{H}_T,T) &= \sum_{i=1}^{N_g(T)} \log 
  \lambda_g^*(t_i) - \int_0^T \lambda_g^*(s) ds + \sum_{i=1}^{N_g(T)} \log f(m) \\
&= \sum_{i=1}^{N_g(T)} \log \left(\lambda + \sum_{j=1}^{i-1} A e^{\alpha m_j} 
  \beta e^{-\beta(t_i-t_j)}\right) - \lambda t - \sum_{i=1}^{N_g(T)} A  
  e^{\alpha m_i} \left(1 - e^{-\beta(t - t_i)}\right) \\
&\quad + \sum_{i=1}^{N_g(T)} \gamma e^{-\gamma m_i}
\end{aligned}
$$

$\gamma$ only appears in the last term, so will be able to estimated 
independently of the other variables. For MLE, we can use the estimator 
$\hat{\gamma}=1/\bar{\gamma}$ and 
$\mathrm{SE}(\hat{\gamma}) = \hat{\gamma}/\sqrt{n}$.

```{r}
loglik <- function(theta, T_max, H_t) {
  # this is an expensive O([N(T_max)]^2) computation
  log_intensities <- sapply(1:nrow(H_t), function(i) {
    t_i <- H_t$time[i]
    past_events <- H_t[H_t$time < t_i, ]
    log(exp_intensity(theta = theta, t = t_i, H_t = past_events))
  })
  
  # total log-likelihood = sum(log(intensities)) - compensator(T)
  sum(log_intensities) - exp_compensator(theta = theta, t = T_max, H_t = H_t)
}

neg_loglik <- function(theta, T_max, H_t) {-loglik(theta, T_max, H_t)}

fit_exp <- optim(
  par = c(1, 1, 1, 1), 
  fn = neg_loglik, 
  H_t = simulated_process,
  T_max = T_max,
  method = "L-BFGS-B",
  lower = c(1e-6, 1e-6, 1e-6, 1e-6),
  control=list(trace=2),
  hessian = TRUE
)

mle_exp <- fit_exp$par
v_cov_exp <- solve(fit_exp$hessian)
se_exp <- sqrt(diag(v_cov_exp))

cat("\nMaximum Likelihood Estimates with 95% CIs:\n", sprintf(
  "lambda_mle = %.3f ± %.3f \n A_mle = %.3f ± %.3f \n alpha_mle = %.3f ± %.3f \n beta_mle = %.3f ± %.3f \n gamma_mle = %.3f ± %.3f \n\n", 
  mle_exp[1], 1.96*se_exp[1], 
  mle_exp[2], 1.96*se_exp[2], 
  mle_exp[3], 1.96*se_exp[3], 
  mle_exp[4], 1.96*se_exp[4],
  1/mean(simulated_process$mark), 1.96*1/(mean(simulated_process$mark)*sqrt(nrow(simulated_process)))
))

cat("True Values = ", theta, "\n")
```

Below is some code for implementing a Gibbs Sampler to sample from the posterior
distributions of the models parameters.

It will be useful to define functions for these frequently computed terms:
$$
\begin{gather}
S_i \coloneqq \sum_{t_j<t_i}e^{\alpha m_j}\beta e^{-\beta(t_i-t_j)} \\
H \coloneqq \sum_{j=1}^n e^{\alpha m_j} e^{-\beta(t-t_j)}
\end{gather}
$$

```{r}
compute_S_H <- function(times, marks, alpha, beta, T_max) {
  
  n <- length(times)
  S <- numeric(n)
  
  for (i in 2:n) {
    dt <- times[i] - times[1:(i-1)]
    S[i] <- beta*sum(exp(alpha*marks[1:(i-1)] - beta*dt))
  }
  
  # compensator part (excluding lambda T)
  H <- sum(exp(alpha*marks) * (1 - exp(-beta*(T_max - times))))
  
  return(list(S = S, H = H))
}
```

Functions for full conditional log-posteriors, where we have used the 
log-transformation to ensure positive values. The full conditional 
log-posteriors will need the jacobian of the transformation. E.g. for sampling 
$\lambda$ on the log-scale:

$$
\tilde{\lambda} = \log \lambda \quad \Rightarrow \quad \lambda=e^{\tilde{\lambda}}
$$
The transformation of the posterior is 
$$
\log p(\tilde{\lambda}\mid\cdot)=\log p(\lambda = e^{\tilde{\lambda}}\mid\cdot) 
  + \log \left|\frac{d\lambda}{d\tilde{\lambda}} \right| 
  = \log p(\lambda = e^{\tilde{\lambda}} \mid \cdot) + \log\lambda
$$


```{r}
# '+ log_lambda' etc on end of each function return is jacobian of transformation

logpost_lambda <- function(log_lambda, A, S, T_max, r_lambda) {
  lambda <- exp(log_lambda)
  sum(log(lambda + A*S)) - lambda*(T_max + r_lambda) + log_lambda  
}

logpost_A <- function(log_A, lambda, S, H, r_A) {
  A <- exp(log_A)
  sum(log(lambda + A*S)) - A*(r_A + H) + log_A 
}

# not sampling log-transformed variable as could be positive or negative
# therefore don't need jacobian
logpost_alpha <- function(alpha, lambda, A, times, marks, beta, T_max, r_alpha) {
  # need to recompute S, H with updated alpha
  tmp <- compute_S_H(times, marks, alpha, beta, T_max)
  S <- tmp$S
  H <- tmp$H
  sum(log(lambda + A*S)) - r_alpha*alpha - A*H
}

logpost_beta <- function(log_beta, lambda, A, times, marks, alpha, T_max, r_beta) {
  beta <- exp(log_beta)
  # need to recompute S, H with updated alpha, beta
  tmp <- compute_S_H(times, marks, alpha, beta, T_max)
  S <- tmp$S
  H <- tmp$H
  
  sum(log(lambda + A*S)) - r_beta*beta - A*H + log_beta
}

# posterior for gamma with gamma prior has gamma posterior
sample_gamma <- function(marks, a0, b0) {
  n <- length(marks)
  shape <- a0 + n
  rate  <- b0 + sum(marks)
  
  rgamma(1, shape = shape, rate = rate)
}

mh_step <- function(current, logpost_fun, proposal_sd, ...) {
  # random walk sampler (in log-space)
  proposal <- rnorm(1, current, proposal_sd)
  
  #log(acceptance ratio)
  log_acc <- logpost_fun(proposal, ...) -
             logpost_fun(current, ...)
  
  if (log(runif(1)) < log_acc) {
    return(list(value = proposal, accept = 1))
  } else {
    return(list(value = current, accept = 0))
  }
}
```

Code for Metropolis-within-Gibbs sampler
 
```{r}
run_sampler <- function(times, marks, T_max, n_iter, init, prior_params, proposal_sds) {
  
  samples <- matrix(NA, n_iter, 5)
  acceptance <- matrix(NA, n_iter, 5)
  colnames(samples) <- c("lambda", "A", "alpha", "beta", "gamma")
  colnames(acceptance) <- c("lambda", "A", "alpha", "beta", "gamma")
  
  # initialise
  lambda <- init$lambda
  A      <- init$A
  alpha  <- init$alpha
  beta   <- init$beta
  gamma  <- init$gamma
  
  pb <- txtProgressBar(min = 0,      # Minimum value of the progress bar
                       max = n_iter, # Maximum value of the progress bar
                       style = 3,    # Progress bar style (also available style = 1 and style = 2)
                       width = 50,   # Progress bar width. Defaults to getOption("width")
                       char = "=")   # Character used to create the bar
  
  for (iter in 1:n_iter) {
    
    # Precompute S and H for Lambda and A sampling
    tmp <- compute_S_H(times, marks, alpha, beta, T_max)
    S <- tmp$S
    H <- tmp$H
    
    # Lambda (Metropolis-within-Gibbs)
    res <- mh_step(log(lambda), logpost_lambda, proposal_sds$lambda,
                   A = A, S = S, T_max = T_max, r_lambda = prior_params$r_lambda)
    lambda <- exp(res$value)
    acceptance[iter, "lambda"] <- res$accept
    
    # A (Metropolis-within-Gibbs)
    res <- mh_step(log(A), logpost_A, proposal_sds$A, lambda = lambda, 
                   S = S, H = H, r_A = prior_params$r_A)
    A <- exp(res$value)
    acceptance[iter, "A"] <- res$accept
    
    # Alpha (Metropolis-within-Gibbs)
    res <- mh_step(alpha, logpost_alpha, proposal_sds$alpha,
                   lambda = lambda, A = A,
                   times = times, marks = marks,
                   beta = beta, T_max = T_max,
                   r_alpha = prior_params$r_alpha)
    alpha <- res$value
    acceptance[iter, "alpha"] <- res$accept
    
    # Beta (Metropolis-within-Gibbs)
    res <- mh_step(log(beta), logpost_beta, proposal_sds$beta,
                   lambda = lambda, A = A,
                   times = times, marks = marks,
                   alpha = alpha, T_max = T_max,
                   r_beta = prior_params$r_beta)
    beta <- exp(res$value)
    acceptance[iter, "beta"] <- res$accept
    
    # Gamma (pure Gibbs)
    gamma <- sample_gamma(marks, prior_params$a_gamma, prior_params$b_gamma)
    acceptance[iter, "gamma"] <- 1
    
    samples[iter, ] <- c(lambda, A, alpha, beta, gamma)
    setTxtProgressBar(pb, iter)
  }
  
  close(pb)
  acceptance_rates <- colMeans(acceptance)
  
  return(list(samples = samples, acceptance_rates = acceptance_rates))
}
```


```{r}
########################################################################
# Uncomment to rerun MCMC chains if ./data/chains.rds doesn't exist 
#########################################################################

# library(parallel)
# 
# times <- simulated_process$time
# marks <- simulated_process$mark
# prior_params <- list(
#   a_gamma = 1, b_gamma = 1,
#   r_lambda = 0.1, r_A = 0.1, r_alpha = 0.1, r_beta = 0.1
# )
# proposal_sds <- list(lambda = 0.25, A = 0.2, alpha = 0.12, beta = 0.2)
# n_chains <- 4
# n_iter   <- 5000
# 
# # reproducible initial values generated randomly to check chain convergence
# set.seed(6)
# init_list <- lapply(1:n_chains, function(i) {
#   list(
#     lambda = runif(1, 0.01, 1),
#     A      = runif(1, 0.01, 1),
#     alpha  = rnorm(1, 0, 1),
#     beta   = runif(1, 0.5, 5),
#     gamma  = runif(1, 0.5, 5)
#   )
# })
# 
# # reproducible parallel MCMC
# cl <- makeCluster(n_chains)
# clusterSetRNGStream(cl, iseed = 123)
# 
# clusterExport(cl, c("times", "marks", "T_max", "n_iter", "init_list", "prior_params",
#                     "proposal_sds", "run_sampler", "mh_step", "compute_S_H",
#                     "logpost_lambda", "logpost_A", "logpost_alpha",
#                     "logpost_beta", "sample_gamma"))
# 
# chains <- parLapply(cl, 1:n_chains, function(i) {
#   run_sampler(
#     times = times,
#     marks = marks,
#     T_max = T_max,
#     n_iter = n_iter,
#     init = init_list[[i]],
#     prior_params = prior_params,
#     proposal_sds = proposal_sds
#   )
# })
# stopCluster(cl)
# saveRDS(chains, "./data/chains.rds")
```

```{r}
library(coda)
library(MCMCvis)

if (file.exists("./data/chains.rds")) {chains <- readRDS("./data/chains.rds")}
mcmc_chains <- lapply(chains, function(chain_res) {
  # chain_res$samples is n_iter x n_params matrix
  mcmc(chain_res$samples)
})

# Combine into mcmc.list
chains_coda <- mcmc.list(mcmc_chains)
chains_coda_thin <- window(chains_coda, start = 1001, thin = 10)

MCMCtrace(chains_coda_thin, , ind = TRUE, Rhat = TRUE, n.eff = TRUE, pdf = FALSE)

# Posterior summary
MCMCsummary(chains_coda_thin, round = 4)
```

